---
layout: post
title: "Better Confidence Intervals for Quantiles"
tags: [rstats, stats, nonparametrics]
bibliography: ~/Literature/Bibtex/jabref.bib
comments: true
---

```{r,include=FALSE,echo=FALSE,message=FALSE}
##If default fig.path, then set it.
if (knitr::opts_chunk$get("fig.path") == "figure/") {
  knitr::opts_knit$set( base.dir = '/Users/hoehle/Sandbox/Blog/')
  knitr::opts_chunk$set(fig.path="figure/source/2016-10-23-quantileCI/")
}
fullFigPath <- paste0(knitr::opts_knit$get("base.dir"),knitr::opts_chunk$get("fig.path"))
filePath <- "/Users/hoehle/Sandbox/Blog/figure/source/2016-10-23-quantileCI/"

knitr::opts_chunk$set(echo = TRUE,fig.width=8,fig.height=5,fig.cap='') # autodep=TRUE
options(width=90)
library("dplyr")
library("ggplot2")
library("tidyr")
library("methods")
theme_set(theme_bw())
```
$$
\newcommand{\bm}[1]{\boldsymbol{\mathbf{#1}}}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
$$

## Abstract

We discuss the computation of confidence intervals for the median or
any other quantile in R. In particular we are interested in the
interpolated order statistic approach suggested by
@hettmansperger_sheather1986 and @nyblom1992. In order to make the
methods available to a greater audience we provide an implementation
of these methods in the R package `quantileCI` and a small simulation
study is conducted to show that these intervals indeed have a very
good coverage. The study also shows that these intervals perform
better than the currently available approaches in R. We therefore
propose that these intervals should be used more in the future!

<center>
```{r,results='asis',echo=FALSE}
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"FIRSTPICTURE-1.png"),")")
```
</center>

{% include license.html %}

## Introduction

Statistics 101 teaches that for a distribution, possibly contaminated with outliers, a robust measure of the central tendency is the median. Not knowing this fact can make your analysis worthy to report in the [newspaper](http://www.sueddeutsche.de/wirtschaft/heilbronn-dieser-mann-ist-so-reich-dass-statistiken-seines-wohnorts-wertlos-sind-1.2705044) ([Google translate](https://translate.google.com/translate?sl=de&tl=en&js=y&prev=_t&hl=en&ie=UTF-8&u=http%3A%2F%2Fwww.sueddeutsche.de%2Fwirtschaft%2Fheilbronn-dieser-mann-ist-so-reich-dass-statistiken-seines-wohnorts-wertlos-sind-1.2705044&edit-text=)).

Higher quantiles of a distribution also have a long history as threshold for when to declare an observation an outlier. For example, growth curves for children illustrate how the quantiles of, e.g., [the BMI distribution develop by age](http://www.cdc.gov/growthcharts/data/set1clinical/cj41l024.pdf). **Obesity** is then for children defined as exceedance of the [97.7% quantile](http://www.who.int/growthref/bmifa_girls_z_5_19_labels.pdf?ua=1) of the distribution at a particular age. Quantile regression is a non-parametric method to compute such curves and the statistical community has been quite busy lately investigating new ways to compute such quantile regressions models.

The focus of this blog post is nevertheless the simplest setting:
Given an iid. sample $\bm{x}$ of size $n$ from a univariate and
absolutely continuous distribution $F$, how does one compute an
estimate for the $p$-Quantile of $F$ together with a corresponding
two-sided $(1-\alpha)\cdot 100\%$ confidence interval for it?

### The Point Estimate

Computing the quantile in a sample with statistical software is discussed in the excellent survey of @hyndman_fan1996. The simplest estimator is based on the [order statistic](https://en.wikipedia.org/wiki/Order_statistic) of the sample, i.e. $x_{(1)} < x_{(2)} < \cdots < x_{(n)}$.
$$
\hat{x}_p = \min_{k} \left\{\hat{F}(x_{(k)}) \geq p\right\} = x_{(\lceil n \cdot p\rceil)},
$$
where $\hat{F}$ is the [empirical cumulative distribution](https://en.wikipedia.org/wiki/Empirical_distribution_function) function of the sample. Since $\hat{F}$ has jumps of size $1/n$ the actual value of $\hat{F}(\hat{x}_{p})$ can end up being somewhat larger than the desired $p$. Therefore,
 @hyndman_fan1996 prefer estimators interpolating between the two values of the order statistic with $\hat{F}$ just below and just above $p$. It is interesting that even [20 years after](http://robjhyndman.com/hyndsight/sample-quantiles-20-years-later/), there still is no universally accepted way to do this in different statistical software and the `type` argument of the `quantile` function in R has been a close friend when comparing results with SPSS or Stata users. In what follows we will, however, stick with the simple $x_{(\lceil n \cdot p\rceil)}$ estimator stated above.

Below is illustrated how one would use R to compute the, say, the 80%
quantile of a sample using the above estimator. We can compute this
either manually or using the `quantile` function with `type=1`:

```{r,FIX_SEED_VALUE,echo=FALSE}
set.seed(as.integer(charToRaw("R")))
```
```{r}
##Make a tiny artificial dataset, say, the BMI z-score of 25 children
sort(x <- rnorm(25))
##Define the quantile we want to consider
p <- 0.8
##Since we know the true distribution we can easily find the true quantile
(x_p <- qnorm(p))
##Compute the estimates using the quantile function and manually
c(quantile=quantile(x, type=1, prob=p), manual=sort(x)[ceiling(length(x)*p)])
```

### Confidence interval for the quantile

Besides the point estimate $\hat{x}_p$ we also would like to report a two-sided $(1-\alpha)\cdot 100\%$ confidence interval $(x_p^{\text{l}}, x_p^{\text{u}})$ for the desired population quantile. The interval $(x_p^{\text{l}}, x_p^{\text{u}})$ should, hence, fulfill the following condition:
$$
P( (x_p^{\text{l}}, x_p^{\text{u}}) \ni x_p) = 1 - \alpha,
$$
where we have used the "backwards" $\in$ to stress the fact that it's the interval which is random. Restricting the limits of this confidence intervals to be **one of the realisations from the order statistics** implies that we need to find indices $d$ and $e$ with $d<e$ s.t.
$$
P( x_{(d)} \leq x_p \leq x_{(e)}) \geq 1 - \alpha.
$$
Note that it may not be possible to achieve the desired coverage exactly in this case. For now we prefer the conservative choice of having to attain **at least** the desired coverage. Note that for $1\leq r \leq n$ we have
$$
\begin{align*}
P( x_{r} \leq x_p) &= P(\text{at least $r$ observations are smaller than or equal to $x_p$}) \\
      &= \sum_{k=r}^{n} P(\text{exactly $k$ observations are smaller than or equal to $x_p$}) \\
      &= \sum_{k=r}^{n} {n \choose k} P(X \leq x_p)^k (1-P(X \leq x_p))^{n-k} \\
      &= \sum_{k=r}^{n} {n \choose k} p^k (1-p)^{n-k} \\
      &= 1 - \sum_{k=0}^{r-1} {n \choose k} p^k (1-p)^{n-k}
\end{align*}
$$
In principle, we could now try out all possible $(d,e)$ combinations and for each interval investigate, whether it has the desired $\geq 1-\alpha/2$ property. If several combinations achieve this criterion we would, e.g., take the interval having minimal length. This is what the `MKmisc::quantileCI` function does. However, the number of pairs to investigate is of order $O(n^2)$, which for large $n$ quickly becomes lengthy to compute. Instead, we compute an **equi-tailed confidence interval** by finding two one-sided $1-\alpha/2$ intervals, i.e. we find $d$ and $e$ s.t. $P(x_{(d)} \leq x_p) = 1-\alpha/2$ and
$P(x_p \geq x_{(e)}) = 1-\alpha/2$. In other words,

$$
\begin{align*}
d &= \argmax P(x_{(r)} \leq x_p) \geq 1 - \frac{\alpha}{2} \\
  &= \texttt{qbinom(alpha/2, size=n, prob=p)} \\
e &= \argmin P(x_p \geq x_{(r)}) \geq 1 - \frac{\alpha}{2} \\
  &= \texttt{qbinom(1-alpha/2, size=n, prob=p) + 1}
\end{align*}
$$

Note that the problem can arise, that the above solutions are zero or
$n+1$, respectively. In this case one has to decide how to
proceed. For an illustration of the above in case of the median see
the [post ](http://freakonometrics.hypotheses.org/4199) by
[\@freakonometrics](https://twitter.com/freakonometrics). Also note
that the `qbinom` function uses the
[Cornish-Fisher Expansion](https://en.wikipedia.org/wiki/Cornish%E2%80%93Fisher_expansion)
to come up with an initial guess for the quantile, which is then
refined by a numerical search. In other words, the function is of
order $O(1)$ and will, hence, be fast even for large $n$.

When it comes to confidence intervals for quantiles the set of alternative implementations in R is extensive. [Searching for this on CRAN](http://finzi.psych.upenn.edu/cgi-bin/namazu.cgi?query=confidence+interval+for+quantiles&max=100&result=normal&sort=score&idxname=functions&idxname=vignettes&idxname=views),  we found the following functionality:

```{r,echo=FALSE}
#From https://www.r-bloggers.com/list-of-user-installed-r-packages-and-their-versions/
ip <- as.data.frame(installed.packages()[,c(1,3:4)])
rownames(ip) <- NULL
ip <- ip[is.na(ip$Priority),1:2,drop=FALSE]

getVersion <- function(pkgName) {
ip %>% filter(Package == pkgName) %>% select(Version) %>% as.matrix %>% as.character
}
```

| Package::Function   |  Version | Description                                            |
|---------------------|:--------:|--------------------------------------------------------|
| [`MKMisc::quantileCI`](http://finzi.psych.upenn.edu/R/library/MKmisc/html/quantileCI.html)| `r getVersion("MKmisc")` | Implements an exact but very slow $O(n^2)$ search as well as an asymptotic method approximating the exact procedure. Do to the method being slow it is not investigated further, but looking at it an `Rcpp` implementation of the nested loop might be able to speed up the performance substantially. |
|   |   |  |
| [`jmuOutlier::quantileCI`](http://finzi.psych.upenn.edu/R/library/jmuOutlier/html/quantileCI.html)  |  `r getVersion("jmuOutlier")`  | Implements the exact method. |
|   |   |  |
| [`envStats::eqnpar`](http://finzi.psych.upenn.edu/R/library/EnvStats/html/eqnpar.html)  | `r getVersion("EnvStats")` | implements both an exact and an asymptotic interval  |
|   |   |  |
| [`asht::quantileTest`](http://finzi.psych.upenn.edu/R/library/asht/html/quantileTest.html)| `r getVersion("asht")` | also implements an exact method |
|   |   |  |
| [`Qtools::confint.midquantile`](http://finzi.psych.upenn.edu/R/library/Qtools/html/confint.midquantile.html) | `r getVersion("Qtools")` | operates on the mid-quantile (whatever that is). The method is not investigated further.
|   |   |  |
<p>

There might even be more, but for now we are satisfied comparing just
the above mentioned procedures:

```{r,results='hold'}
as.numeric(MKmisc::quantileCI(x=x, prob=p, method="exact",conf.level=0.95)$CI)
as.numeric(MKmisc::quantileCI(x=x, prob=p, method="asymptotic",conf.level=0.95)$CI)
as.numeric(jmuOutlier::quantileCI(x=x, probs=p, conf.level=0.95)[1,c("lower","upper")])
as.numeric(EnvStats::eqnpar(x=x, p=p, ci=TRUE, ci.method="exact",approx.conf.level=0.95)$interval$limits)
as.numeric(EnvStats::eqnpar(x=x, p=p, ci=TRUE, ci.method="normal.approx",approx.conf.level=0.95)$interval$limits)
as.numeric(asht::quantileTest(x=x,p=p,conf.level=0.95)$conf.int)
```

An impressive number of similar, but yet, different results! To add to the confusion here is
our take at this as developed in the `quantileCI` package available from github:

```{r,eval=FALSE}
devtools::install_github("hoehleatsu/quantileCI")
```

The package provides three methods for computing confidence intervals for quantiles:
```{r, results='hold'}
quantileCI::quantile_confint_nyblom(x=x, p=p, conf.level=0.95,interpolate=FALSE)
quantileCI::quantile_confint_nyblom(x=x, p=p, conf.level=0.95,interpolate=TRUE)
quantileCI::quantile_confint_boot(x, p=p, conf.level=0.95,R=999, type=1)
```

The first procedure with `interpolate=FALSE` implements the previously
explained exact approach, which is also implemented in some of the
other packages. However, when the `interpolate` argument is set to
`TRUE` (the default), an additional interpolation step between the two
neighbouring order statistics is performed as suggested in the work of
@nyblom1992, which extends work for the median by
@hettmansperger_sheather1986 and generates intervals of the form.

$$
\left( (1-\lambda_1) x_{(d)} + \lambda_1 x_{(d+1)}, (1-\lambda_2) x_{(e-1)} + \lambda_1 x_{(e)} \right)
$$

with $0 \leq \lambda_1, \lambda_2 \leq 1$ chosen appropriately to get
as close to the desired coverage as possible without knowing the
exact underlying distribution - see the paper for details.

The last call in the above is to a basic bootstrap procedure, which
resamples the data with replacement, computes the quantile using
`type=1` and then reports the 2.5% and 97.5% percentiles of this
bootstrapped distribution. Such percentiles of the basic bootstrap are
a popular way to get confidence intervals for the quantile, e.g., this
is what we have used in @hoehle_hoehle2009 for reporting the 95%
quantile of the absolute difference in height at so called **check
points** in the assessment of accuracy for a digital elevation model
(DEM) in photogrammetry. However, the coverage of the percentile
bootstrap procedure is not without problems, because the convergence
rate as a function of the number of replicates $r$ is only of order
$O(r^{-\frac{1}{2}})$ for quantiles (@falk_kaufmann1991).  As a
trade-off between accuracy and speed we use $R=999$ throughout this
post.

## Simulation Study to Determine Coverage

```{r,echo=FALSE}
##Function to compute different methods on same x.
quantile_confints <- function(x, p, conf.level, x_is_sorted=FALSE) {
  if (!x_is_sorted) { x <- sort(x)}

  ##Compute the various confidence intervals as above
  res <- data.frame(jmuOutlier_exact=as.numeric(jmuOutlier::quantileCI(x=x, probs=p, conf.level=conf.level)[1,c("lower","upper")]),
                    EnvStats_exact=as.numeric(EnvStats::eqnpar(x=x, p=p, ci=TRUE, ci.method="exact",approx.conf.level=conf.level)$interval$limits),
                    EnvStats_asymp=as.numeric(EnvStats::eqnpar(x=x, p=p, ci=TRUE, ci.method="normal.approx",approx.conf.level=conf.level)$interval$limits),
                    asht_quantTest=as.numeric(asht::quantileTest(x=x,p=p,conf.level=conf.level)$conf.int),
                    nyblom_exact=quantileCI::quantile_confint_nyblom(x=x, p=p, conf.level=conf.level,x_is_sorted=TRUE,interpolate=FALSE),
                    nyblom_interp=quantileCI::quantile_confint_nyblom(x=x, p=p, conf.level=conf.level,x_is_sorted=TRUE,interpolate=TRUE),
                    boot=quantileCI::quantile_confint_boot(x, p=p, conf.level=conf.level,R=999)
  )
  if (p == 0.5) {
    res$hs_interp = quantileCI::median_confint_hs(x=x,  conf.level=conf.level,x_is_sorted=TRUE,interpolate=TRUE)
  }
  return(res)
}
```

We write a function, which for a given sample `x` computes two-sided confidence intervals for the p-Quantile using a selection of the above described procedures:
```{r}
quantile_confints(x, p=p, conf.level=0.95)
```

In order to evaluate the various methods and implementations we
conduct a Monte Carlo simulation study to assess each methods'
coverage. For this purpose we write a small wrapper function to
conduct the simulation study using
[parallel computation](http://gforge.se/2015/02/how-to-go-parallel-in-r-basics-tips/). The
function wraps the `quantileCI::qci_coverage_one_sim` function,
which lets the user define a simulation scenario (true underlying
distribution, size of the sample, etc.), then applies all confidence
interval methods gathered in the above `quantile_confints`
and finally assesses whether each confidence interval covers the true
value or not.

```{r}
simulate.coverage_qci <- function(n=n,p=p,conf.level=0.9, nSim=10e3, ...) {
  ##Windows users: change to below lapply function or use snow.
  ##lapplyFun <- function(x, mc.cores=NA, ...) pblapply(x, ...)
  lapplyFun <- parallel::mclapply

  sims <- dplyr::bind_rows(
    lapplyFun(1L:nSim, function(i) {
      quantileCI::qci_coverage_one_sim(qci_fun=quantile_confints, n=n,p=p,conf.level=conf.level,...)
    }, mc.cores = parallel::detectCores() - 1)
  ) %>% summarise_each(funs(mean))
  return(sims)
}
```

#### Simulation 1
We can now compare the coverage of the different implementation for the particular `n`=25 and `p`=`r p` setting:
```{r, SIMSTUDY_N25_p80, cache=TRUE}
simulate.coverage_qci(n=25, p=0.8, conf.level=0.95)
```

Note that the `nyblom_interp` procedure is closer to the nominal
coverage than it's exact cousin `nyblom_exact` and the worst results
are obtained by the bootstrap percentile method.  We also note that
the results of the `jmuOutlier_exact` method appear to deviate from
`asht_quantTest` as well as `nyblom_exact`, which is surprising,
because they should implement the same approach.

#### Simulation 2
As a further test-case we consider the situation for the median in a
smaller sample:
```{r, SIMSTUDY_N11_MEDIAN, cache=TRUE}
simulate.coverage_qci(n=11, p=0.5, conf.level=0.95)
```

We note that the `EnvStats_exact` procedure again has a lower coverage
than the nominal required level, it must therefore implement a
slightly different procedure than expected. That coverage is
less than the nominal for an *exact* method is, however, somewhat
*surprising*.

In this study for the median, the original
@hettmansperger_sheather1986 procedure implemented in `quantileCI` as
function `median_confint_hs` is also included in the comparison
(`hs_interp`).  Note that the @nyblom1992 procedure for $p=\frac{1}{2}$ just
boils down to this approach. Since the neighbouring order statistics
are combined using a weighted mean, the actual level is just close to
the nominal level. It can, as observed for the above setting, be
slightly lower than the nominal level. The bootstrap method again
doesn't look too impressive.

#### Simulation 3
We finally also add one of the scenarios from Table 1 of the
@nyblom1992 paper, which allows us to check our implementation against
the numerical integration performed in the paper to assess coverage.

```{r, SIMSTUDY_N11_p25, cache=TRUE}
simulate.coverage_qci(n=11, p=0.25, conf.level=0.90)
```

In particular the results of `EnvStats_exact` look disturbing. The
coverage of the interpolated order statistic approach again looks
convincing.

#### Simulation 4

Finally, a setup with a large sample, but now with the t-distribution
with one degree of freedom:

```{r, SIMSTUDY_N101_MEDIAN, cache=TRUE}
simulate.coverage_qci(n=101, p=0.9, rfunc=rt, qfunc=qt, conf.level=0.95, df=1)
```
Again the interpolation method provides the most convincing
results. The bootstrap on the other hand again has the worst coverage,
a larger $R$ might have helped here, but would have made the
simulation study even more time consuming.

# Conclusion and Future Work

Can we based on the above recommend one procedure to use in practice?
Well, even though the simulation study is small, the exact
`EnvStats::eqnpar` approach appears to yield below nominal coverage
intervals, sometimes even substantially, and hence is not to be
recommended. On the other hand, `jmuOutlier_exact`, `asht_quantTest`,
and `nyblom_exact` in all four cases provide above nominal level
coverage, i.e. the intervals are conservative in as much as they are
too wide. Also slightly disturbing is that the results of the exact
confidence interval methods varied somewhat between the different R
implementations. Part of the differences arise from handling the
discreteness of the procedure as well as the edge cases
differently. The basic percentile bootstrap method is a simple
approach, providing acceptable, but not optimal coverage and also
depends in part on $R$. In particular for very large $n$ or for a
large number of replication in the simulation study, the method with a
large $R$ can be slow. Suggestions exist in the literature on
how to improve the speed of coverage convergence by smoothing (see,
e.g., @deangelis_etal1993), but such work is beyond the scope of this
post.

The @hettmansperger_sheather1986 and @nyblom1992 method, respectively,
provide very good coverage close to the nominal level. The method is
fast to compute, available through the `quantileCI` R package and
would be our recommendation to use in practice.

```{r,echo=FALSE}
##Get back the old data from the first example in the post
set.seed(as.integer(charToRaw("R")))
n <- 25 ; x <- sort(rnorm(n)) ; p <- 0.8 ; alpha <- 0.1
x_p <- qnorm(p)
hat_x_p <- quantile(x, prob=p, type=1)
```

Altogether, we summarise our findings as follows: **More confidence in
confidence intervals for quantiles!** and let the following picture
illustrating 90% confidence intervals for the `r p*100`%
quantile of the standard normal distribution based on the above sample
of size $n$=`r n` say this in less than 1000 words.

```{r,PREPAREPICTURE,echo=FALSE,results='hide'}
ci1 <- quantileCI::quantile_confint_nyblom(x, p=p, interpolate=FALSE, conf.level=1-alpha)
ci2 <- quantileCI::quantile_confint_nyblom(x, p=p, interpolate=TRUE, conf.level=1-alpha)

idx <- pmatch(ci1, x)
#Tail probabilities
(p_tail <- c(pbinom(idx[1] - 1, prob=p, size=n),   1-pbinom(idx[2]-1, prob=p, size=n)))
cov1 <- 1-sum(p_tail)
ci3 <- x[c(idx[1]+1,idx[2]-1)]
cov3 <- 1 - sum(c(pbinom(idx[1] + 1 - 1, prob=p, size=n),   1-pbinom(idx[2]-1-1, prob=p, size=n)))

#Compute coverage for smoothed interval by Monte Carlo sampling (this might be slow)
#Try to do this using numerical integration instead.
tails  <- apply(replicate(1e6, {
  y <- sort(rnorm(n))
  ##These numbers are computed by the Nyblom procedure. Computed manually
  ##from the functions defined inside the quantile_confint_nyblom function.
  d <- 17 ; e <- 24
  lambda1 <- 0.09313647
  lambda2 <- 0.4259059
  c((1-lambda1)*y[d] + lambda1*y[d+1] > x_p, (1-lambda2)*y[e-1] + lambda2*y[e] <= x_p)
}),1,mean)
cov2 <- 1 - sum(tails)
```

```{r,FIRSTPICTURE,fig.height=4,fig.width=8,echo=FALSE,results='hide',fig.align='center'}
par(mar=c(4,2,3,2))
x_extreme <- 1.1*max(abs(x))
x_grid <- seq(-x_extreme,x_extreme,length=1000)

makePlot <- function() {
  plot(x_grid, dnorm(x_grid),type="l", col="steelblue",xlab="x",ylab="",yaxt="n", main=paste0("Estimating the ",p*100,"% quantile of the standard normal with a sample of size n=",n))
  rug(x,lwd=2)
  lines( rep(x_p,2), c(-1e99,1e99), col="gray",lty=2,lwd=2)
  axis(1,x_p,label=substitute(x[foo],list(foo=paste0(p*100,"%"))),las=2)
  axis(1,hat_x_p,label=substitute(hat(x)[foo],list(foo=paste0(p*100,"%"))),las=2)

  showCI <- function(ci, at, lwd=2) {
  	  points( hat_x_p, at, pch=20)
	  lines( ci, rep(at,2), lwd=lwd)
      for (i in 1:2) { lines( rep(ci[i],2), c(at - 0.01,at + 0.01),lwd=lwd) }
	  invisible()
  }

  ##Add text
  text(-1.2,0.1,sprintf("Exact %0.f%% CI (%.1f%% coverage)",(1-alpha)*100,cov1*100)) ; showCI(ci1,at=0.1)
  text(-1.2,0.3,sprintf("CI based on next pair (%.1f%% coverage)",cov3*100)) ; showCI(ci3,at=0.3)
  text(-1.2,0.2,sprintf("Nyblom %0.f%% CI (%.1f%% coverage)",(1-alpha)*100,cov2*100)) ; showCI(ci2,at=0.2)

  thinLineCI <- function(ci) {
   lines( rep(ci[1],2), c(0.1,0.3), lty=3, col="lightgray")
   lines( rep(ci[2],2), c(0.1,0.3), lty=3, col="lightgray")
  }
  #Show thin lines between the two CIs based on the order statistic
  thinLineCI(ci1)
  thinLineCI(ci3)

  invisible()
}

makePlot()
```

# Appendix

Given PDF $f$ and CDF $F$ of the underlying distribution, the coverage
probability of the one sided Nyblom $(1-\alpha/2)\cdot 100\%$
confidence interval $(1-\lambda) x_{(r)} + \lambda x_{(r+1)}$ for
$x_p$ can be found as follows: Let $z = (1-\lambda) x_{(r)} + \lambda
x_{(r+1)}$. If the considered interval is a lower-confidence interval,
we are interested in finding $P( z \leq x_p)=\int_{-\infty}^{x_p}
f_z(z) dz$. Here, the PDF of $z$ is found as

$$
f_z(z) = \int_{-\infty}^{z} f_{x_{(r)},x_{(r+1)}}( x_{(r)},
x_{(r+1)}^*) dx_{(r)}, \quad\text{where}\quad x_{(r+1)}^* = \frac{z-(1-\lambda)x_{(r)}}{\lambda}.
$$

The
[joint distribution of the order statistic](https://en.wikipedia.org/wiki/Order_statistic#The_joint_distribution_of_the_order_statistics_of_an_absolutely_continuous_distribution)
is here

$$
f_{x_{(r)},x_{(r+1)}}( x_{(r)},
x_{(r+1)}) = \frac{n!}{(r-1)!(n-r-1)!} F(x_{(r)})^{r-1}
\left( 1- F(x_{(r+1)})\right)^{n-r-1} f(x_{(r+1)}) f(x_{(r)}).
$$
The two nested integrals can be solved by numerical integration using,
e.g., the `integral` function.

# References


