---
layout: post
title: "suRprise! - Classifying Kinder Eggs by Boosting"
tags: [rstats, stats, programming, juleforsøg, classification]
bibliography: ~/Literature/Bibtex/jabref.bib
comments: true
---

```{r,include=FALSE,echo=FALSE,message=FALSE}
##If default fig.path, then set it.
if (knitr::opts_chunk$get("fig.path") == "figure/") {
  knitr::opts_knit$set( base.dir = '/Users/hoehle/Sandbox/Blog/')
  knitr::opts_chunk$set(fig.path="figure/source/2016-12-23-surprise/")
}
fullFigPath <- paste0(knitr::opts_knit$get("base.dir"),knitr::opts_chunk$get("fig.path"))
filePath <- "/Users/hoehle/Sandbox/Blog/figure/source/2016-12-23-surprise/"

knitr::opts_chunk$set(echo = TRUE,fig.width=8,fig.height=5,fig.cap='',autodep=TRUE)
options(width=90)
library("dplyr")
library("ggplot2")
library("tidyr")
library("methods")
library("forcats")

theme_set(theme_bw())

##Set seed
set.seed(123)
```

## Abstract

Carrying the Danish tradition of Juleforsøg to the realm of
statistics, we use R to classify the figure content of Kinder Eggs
using boosted regression trees for the egg's weight and possible
rattling noises.

<center>
```{r,results='asis',echo=FALSE}
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"pics/figures.jpg"),")")
```
</center>

{% include license.html %}

## Introduction

A **juleforsøg** is the kind of
[exploding experiment](https://www.youtube.com/watch?v=sinQ06YzbJI8)
happening in the last physics or chemistry class before the Christmas
vacation. Not seldomly the teacher, with a look of secrecy, initializes
the class by locking the door mumbling something like "the headmaster
better not see this...". With Christmas approaching fast, here is an
attempt to create a statistical juleforsøg concluding the *Theory
meets practice* 2016 posting season:

The advertisement campaign of the
[Kinder Surprise Eggs](https://en.wikipedia.org/wiki/Kinder_Surprise)
aka. [Kinder Joy](https://en.wikipedia.org/wiki/Kinder_Joy) claims
that the content of every 7th egg is a figure (see
[example](https://blog.kalaydo.de/blog/wp-content/uploads/2016/05/Biene-Maja.jpg)) -
otherwise they contain toys or puzzles, which positively can be
described as junk. Figures, in particular completed series, on the
other hand, can achieve high
[trading values](https://translate.google.com/translate?sl=de&tl=en&js=y&prev=_t&hl=en&ie=UTF-8&u=https%3A%2F%2Fwww.kalaydo.de%2Fblog%2Fwertvolle-ue-ei-figuren%2F&edit-text=&act=url). The
clear goal is thus to optimize your egg hunting strategy in order to
maximize figure content.


## The problem

Your budget is limited, so the question is which egg to select when
standing in the supermarket?

<center>
```{r,results='asis',echo=FALSE}
##Photo taken 17 Oct 2013 in Coop supermarked Södermalm, Stockholm.
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"pics/inshopwithprice.jpg"),")")
```
</center>
<p>

### Various egg selection strategies

It goes without saying that brute force purchasing strategies would be
insane. Hence, a number of egg selection strategies can be observed in
real life:

* The no clue egg enthusiast: Selects an egg at random. With a certain
    probability (determined by the producer and the cleverness of the
    previous supermarked visitors) the egg contains a figure

* The egg junkie: knows a good
  [radiologist](https://www.radiologycafe.com/blog/easter-egg-xray)

* The egg nerd: using
[scale, rattling noises and the barcode](https://translate.google.com/translate?sl=de&tl=en&js=y&prev=_t&hl=en&ie=UTF-8&u=http%3A%2F%2Fwww.eierwiki.de%2Findex.php%3Ftitle%3DTipps_%2526_Tricks_beim_Eierkauf&edit-text=&act=url)
he/she quickly determines whether there is a figure in the egg

We shall in this post be interested in **the statistician's egg
selection approach**: Egg classification based on weight and rattling
noise using 'top-notch' machine learning algorithms - in our case
based on boosted classification trees.


## Data Collection

```{r,echo=FALSE}
#Load data
surprise <- read.table(paste0(filePath,"surprise.txt"),header=TRUE,skip=9)
#Some data munging to get factor variable
surprise <- surprise %>% mutate(rattles=1-rattles_like_figure, #solve ambiguity in the rattling covar
                                rattles_fac = fct_recode(factor(rattles),
                                                         "no"="0",
                                                         "yes"="1"),
                                figure_fac = fct_recode(factor(figure),
                                                     "no" = "0",
                                                     "yes" = "1"))
```
We collected n=`r nrow(surprise)` eggs of which
`r sprintf('%.1f%%',100*mean(surprise$figure))` were figures - the
data are available under a GPL v3.0 license from [github](https://github.com/hoehleatsu/hoehleatsu.github.io/blob/master/figure/source/2016-12-23-surprise/surprise.txt). For
each egg, we determined its **weight** as well as the sound it produced
when being shaken. If the sounds could be characterized as **rattling**
(aka. clattering) this was indicative of the content consisting of
many parts and, hence, unlikely to be a figure.

<center>
```{r,results='asis',echo=FALSE}
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"pics/weightandrattle.jpg"),")")
```
</center>
<p>
Altogether, the first couple of rows of the dataset look as follows.

```{r}
head(surprise, n=5)
```

### Descriptive Data Analysis
```{r,echo=FALSE}
tab <- with(surprise, table(figure, rattles))
```

The fraction of figures in the dataset was
`r sum(surprise$figure)`/`r nrow(surprise)`, which is way higher than
the proclaimed 1/7; possibly, because professionals egg collectors were at work...

Of the `r sum(tab)` analysed eggs, `r colSums(tab)[1]` were categorized as non-rattling.
The probability of such a non-rattling egg really containing
a figure was `r sprintf("%.1f%%",tab[1,1]/(tab[1,1]+tab[2,1])*100)`.
This proportion is not impressive, but could be due to the data
collector's having a different understanding of exactly how the variable
*rattling* was to be interpreted: Does it *rattle*, or does it
*rattle like a figure*? In hindsight, a clearer definition and
communication of this variable would have prevented ambiguity in the
collection.

A descriptive plot of the weight distribution of eggs with and without
figure content shows, that eggs with figures tend to be slightly
heavier:

```{r, WEIGHTPLOT, echo=FALSE, fig.align='center'}
##Plot boxplot & barchart in one figure
p1 <- ggplot(surprise, aes(x=figure_fac, y= weight)) + geom_boxplot() + xlab("") + ylab("Weight (g)") + ggtitle("Figure inside?")
p2 <- ggplot(surprise, aes(x=weight,fill=figure_fac)) + geom_bar() + xlab("Weight (g)") + ylab("Eggs") + scale_fill_manual(name="Figure inside?",values=c("yes"="#66C2A5","no"="#FC8D62")) + ggtitle("")
gridExtra::grid.arrange(p1,p2,ncol=2)
```

Below the proportion (in %) of eggs with figure content per observed weight:
```{r,echo=FALSE}
tabw <- with(surprise, table(weight, figure_fac))
tabw <- tabw / rowSums(tabw)
print(t(tabw*100),digits=3)
```

A simple selection rule based on weight would be to weigh eggs until
you hit a 40g egg. A slightly less certain stopping rule would be to
pick 34g eggs. However, modern statistics is more than counting and
analysing proportions!

## Machine Learning the Egg Content

We use machine learning algorithms to solve the binary classification
problem at hand. In particular, we use the `caret` package [@caret]
and classify figure content using boosted regression trees as implemented in the
[`xgboost`](https://en.wikipedia.org/wiki/Xgboost) package
[@xgboost]. Details on how to use the `caret` package can, e.g.,
be found in the following
[tutorial](https://topepo.github.io/caret/index.html).

```{r,message=FALSE,warning=FALSE,cache=TRUE, results='hide'}
library(caret)

##Grid with xgboost hyperparameters
xgb_hyperparam_grid = expand.grid(
  nrounds = c(25, 100, 1000),
  eta = c(0.01, 0.001, 0.0001),
  max_depth = c(2, 4, 8, 10, 16, 32),
  gamma = 1,
  colsample_bytree = 0.8, min_child_weight = 1, subsample = 0.5
)
##caret training control object
control <- trainControl(method="repeatedcv", number=8, repeats=8, classProbs=TRUE,
                        summaryFunction = twoClassSummary, allowParallel=TRUE)
##train away and do it parallelized...
library(doMC)
registerDoMC(cores = 3)
m_xgb <- train( figure_fac ~ weight * rattles_fac, data=surprise, method="xgbTree",
               trControl=control, verbose=FALSE, metric="ROC", tuneGrid=xgb_hyperparam_grid)
##look at the result
m_xgb
```
```{r,echo=FALSE}
str <- capture.output(m_xgb)
#Skip some lines of the cv output
idx1 <- grep("Resampling results across tuning",str)
idx2 <- grep("Tuning parameter 'gamma' was ",str)
#Reduce output
removeLines <- (idx1+6):(idx2-4)
str[removeLines[1]] <- "  ...  ...        ..."
cat(paste(str[-removeLines[-1]],"\n"))
```
```{r,eval=FALSE, echo=FALSE}
m_xgb$bestTune
sprintf("%.2f",mean(m_xgb$resample$ROC))
```

The average AUC for the `r length(m_xgb$resample$ROC)` resamples is
`r sprintf("%.2f",mean(m_xgb$resample$ROC))`. Average sensitivity and
specificity are `r sprintf("%.1f%%",100*mean(m_xgb$resample$Sens))`
and `r sprintf("%.1f%%",100*mean(m_xgb$resample$Spec))`, respectively.
This shows that predicting figure content with the available data is
better than simply picking an egg at random,
but no figure-guaranteeing strategy appears possible on a per-egg basis.

### Predicting the Content of a Particular Egg

Suppose the egg you look at weighs 36g and when shaken it sounds
like a lot of small parts being moved. In other words:

```{r, echo=TRUE,message=FALSE,warning=FALSE}
predict(m_xgb, newdata = data.frame(weight=36, rattles_fac="yes"),type="prob")
```

Hence, despite the rattling noises, the classifier is of the opinion
that it's slightly more likely than 50% that there is a figure
inside. However, when we opened this particular egg:

<center>
```{r,results='asis',echo=FALSE}
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"pics/car.jpg"),")")
```
</center>
<p>

...a car. Definitely not a figure! The proof of concept disappointment was,
however, quickly counteracted by the surrounding chocolate...

As a standard operating procedure for your optimized future
supermarket hunt, below are shown the classifier's predicted
probabilities for figure content as a function of egg weight and the
`rattles_fac` variable.

```{r,echo=FALSE, CLASSIFIEROUTPUT,fig.align="center"}
nd <- data.frame(weight=rep(30:40,times=2), rattles_fac=rep(c("no","yes"),each=11))
nd$p <- predict(m_xgb, newdata = nd,type="prob")[,2]
ggplot(nd, aes(x=weight, y=p)) + geom_line() + facet_grid(. ~ rattles_fac) + ylab("Probability of figure") + xlab("Weight (g)") + scale_y_continuous(labels=scales::percent)
```

## Discussion

The above post only discusses the optimal selection on a per-egg
basis. One could weight & shake several eggs and then select the one
with the highest predicted probability for containing a figure. Future
research is needed to solve this sequential decision making problem in
an
[optimal way](http://staff.math.su.se/hoehle/blog/2016/06/12/optimalChoice.html).

### Outlook

We have retained a validation sample of 10 eggs and are willing to
send an unconsumed 11th element of the sample to whoever obtains the
best score on this validation sample. Anyone who knows how to upload
this to [kaggle](https://www.kaggle.com)?

<center>
We wish all readers *God jul* and a happy new year!
</center>
<p>

## Acknowledgments

Thanks to former colleagues at the Department of Statistics,
University of Munich, as well as numerous statistics students in
Munich and Stockholm, for contributing to the data collection. In
particular we thank Alexander Jerak for his idea of optimizing figure
hunting in a data driven way more than 10 years ago.

## Literature
