---
layout: post
title: "The Olympic Medal Table Visualized Gapminder Style"
tags: [datascience, rstats, olympic games]
bibliography: ~/Literature/Bibtex/jabref.bib
comments: true
---

```{r,include=FALSE,echo=FALSE,message=FALSE}
##If default fig.path, then set it.
if (knitr::opts_chunk$get("fig.path") == "figure/") {
  knitr::opts_knit$set( base.dir = '/Users/hoehle/Sandbox/Blog/')
  knitr::opts_chunk$set(fig.path="figure/source/2016-08-21-gapMedal/")
}
fullFigPath <- paste0(knitr::opts_knit$get("base.dir"),knitr::opts_chunk$get("fig.path"))
knitr::opts_chunk$set(echo = TRUE,fig.width=8,fig.height=5,fig.cap='')
options(width=90)
library("dplyr")
library("ggplot2")
library("tidyr")
library("rvest")
library("methods")
theme_set(theme_bw())
```

{% include license.html %}

## Abstract

Following Hans Rosling's Gapminder animation style we visualize the total number of medals a country wins during each olympic summer games in relation to the country's [gross domestic product](https://en.wikipedia.org/wiki/Gross_domestic_product) (GDP) per capita. We illustrate how R's data wrangling capabilities provide a useful toolbox to make such an analysis happen.


# Introduction

Long Swedish winter nights are best spent watching [Hans Rosling](https://en.wikipedia.org/wiki/Hans_Rosling)'s inspiring [TED talks](https://www.youtube.com/watch?v=hVimVzgtD6w). Such visualizations help the statistician make points about temporal trends in a x-axis to y-axis relationship, which otherwise might drown in modelling details.
Recently, I stumbled over a [blog post](https://rpubs.com/sjackman/gapminder-gganimate) on how to use the  [`gganimate`](https://github.com/dgrtwo/gganimate) R package to animate the Gapminder data available from the `gapminder` package. In order to perform a similar *Rosling style* animation consider the following: Today, the Olympic Summer Games in Rio de Janeiro end. As usual this spawns a debate, whether the nation's participation has been successful. For this purpose the [olympic medal table](https://en.wikipedia.org/wiki/Olympic_medal_table) is often taken as basis for comparisons, e.g., to mock your [neighbouring countries](http://politiken.dk/sport/ol/ECE3349634/danmark-og-sverige-kaemper-til-det-sidste---men-hvor-daelen-er-norge-henne/). Recent analyses and visualization have been interested in how to correct these tables for, e.g., population size or, more interesting, analyse the influence of GDP. For example:

* Google provides [alternative Olympics medal tables](https://landing.google.com/altmedaltable/)
* Time Magazine discusses whether it is fair [to rank countries by medals achieved alone](http://time.com/4452128/olympics-medals-per-capita-rankings/)

The aim of the present blog note is to visualize how countries perform in the medal table in relation to their GDP per capita. From a technical viewpoint we experiment with using R to scrape the olympic medal tables from Wikipedia and animate the results Gapminder style. **Disclaimer**: We only show the potential of such an analysis and, hence, worry less about the scientific validity of the analysis.

# Data

We use the data of [Gapminder](https://www.gapminder.org/) in order to obtain country specific population and GDP per capita data for each of the years in the period of 1960-2016. The olympic medal tables are 'harvested' from Wikipedia.

## Olympic medal tables

Olympic medal tables were extracted using the `rvest` package from the corresponding Wikipedia pages by using table-extracting-code described in the post by [Cory Nissen](http://blog.corynissen.com/2015/01/using-rvest-to-scrape-html-table.html). The Wikipedia tables contain the current state of the medal table and hence take changes in the medal distribution, e.g. deprivation due to doping, into account. For details on such a table, see for example the [medal table of the 2012 summer games](https://en.wikipedia.org/wiki/2012_Summer_Olympics_medal_table) in London. In order to stay focused we hide
the scraping functionality in the function `scrape_medaltab` - see the code on GitHub for more details.

```{r,SCRAPE,echo=FALSE}
#Base URL of the wikipedia pages
base_url <- "https://en.wikipedia.org/wiki/1960_Summer_Olympics_medal_table"

#############################################################
# Function to scrape medal table of one particular olympic
# summer game from Wikipedia.
#
# Parameters:
#  year - the year of the summer Olympics to scrape
#
# Returns:
#  A data.frame containing the medal table.
#############################################################
scrape_medaltab <- function(year) {
  tableNumber <- ifelse(year %in% c(1960,1984,1992,1996,2000,2004,2008,2012,2016), 2,1) #table number depends on year
  medals <- gsub("1960",year,base_url) %>%
    read_html() %>%
    html_nodes(xpath=paste0("//*[@id=\"mw-content-text\"]/table[",tableNumber,"]")) %>%
    html_table(fill=TRUE) #%>% .[[1]]
  medals <- medals[[1]]
  #Sometimes the nation is called NOC, and "Rank" is also " Rank " once
  names(medals) <- gsub("NOC","Nation",names(medals))
  names(medals) <- gsub("\\sRank\\s","Rank",names(medals))

  #Remove total row
  medals <- medals %>% filter(row_number() < nrow(medals))

  # Massage country names
  medals <- medals %>% mutate(Nation = gsub("\\*", "", Nation)) #host nation
  medals <- medals %>% mutate(Nation = gsub("‡", "", Nation))   #changes in medals
  medals <- medals %>% mutate(Nation = gsub("^\\s", "", Nation))

  medals <- medals %>% mutate(Nation = gsub("\\([A-Z\\s]+\\)$", "", Nation))
  medals <- medals %>% mutate(Nation = gsub("\\s$", "", Nation))

  return(cbind(Year=year,medals))
}
```

```{r}
#Years which had olympic games
olympic_years <- seq(1960, 2016, by=4)

# Extract olympic medal table from all olympic years since 1960
medals <- bind_rows(lapply(olympic_years, scrape_medaltab))

# Show result
DT::datatable(medals)
```

## Gapminder data

We obtain GDP per capita and population data from [Gapminder](https://www.gapminder.org/data/). Unfortunately, these need to be fetched and merged manually. A more convenient way would have been to take these directly from the package [`gapminder`](https://cran.r-project.org/web/packages/gapminder/index.html), but newer [GDP data](https://www.gapminder.org/data/documentation/gd001/) are now available. Again, we hide the details of the data wrangling activities and refer to GitHub code.

```{r,echo=FALSE, warning=FALSE}
##Library to read excel files
library("openxlsx")

# Obtain GDP/capita data from
# https://www.gapminder.org/data/documentation/gd001/
# From the file: 	Gross Domestic Product per capita by Purchasing
# Power Parities (in international dollars, fixed 2005 prices).
# The inflation and differences in the cost of living between
# countries has been taken into account.

if (!file.exists("../downloads/gapdata001_v14.xlsx")) {
   download.file(url="https://www.gapminder.org/documentation/documentation/gapdata001%20v14.xlsx",destfile="../downloads/gapdata001_v14.xlsx")
}

#Load data from downloaded Excel file and rename columns
gdp_long <- read.xlsx("../downloads/gapdata001_v14.xlsx", sheet=2)
names(gdp_long)[c(1,3)] <- c("Nation","GDPpc")
names(gdp_long)[14:16] <- paste0(names(gdp_long)[14:16],".",1:3)

#Reshaping from wide format to long format
gdp_long <- gdp_long %>% filter(Year >= 1950 & !is.na(GDPpc)) %>% select(Nation, Year, GDPpc)


if (!file.exists("../downloads/gapminder_population.xlsx")) {
  download.file(url="https://spreadsheets.google.com/pub?key=phAwcNAVuyj0XOoBL_n5tAQ&output=xls",destfile="../downloads/gapminder_population.xlsx")
}
pop <- read.xlsx("../downloads/gapminder_population.xlsx")
names(pop) <- gsub("\\.0","",names(pop))
names(pop)[1] <- c("Nation")
pop_long <- pop %>% gather(Year, Population, -Nation)
pop_long <- pop_long %>% mutate(Year=as.numeric(Year))

#Carry last value forward to get population size for 2016
pop_long <- pop_long %>% bind_rows(pop_long %>% filter(Year == 2015) %>% mutate(Year = 2016))

#All this is part of the gapminder package, but unfortunately with uneven recordings
gapminder_manual <- inner_join(gdp_long, pop_long, by=c("Nation","Year"))
```

For convenience, we also extract the corresponding continent each country belongs to. This can be done conveniently by comparing with the `gapminder` dataset (see code for details).

```{r,echo=FALSE,warning=FALSE}
#Fetch continent from gapminder package
library("gapminder")
data("gapminder_unfiltered")
gapminder_continent <- gapminder_unfiltered %>% rename(Continent = continent, Nation = country) %>% select(Nation,Continent) %>% distinct
gapminder_manual <- left_join(gapminder_manual, gapminder_continent, by=c("Nation"))
```

## Joining the two data sources

In principle, all that is left to do is to join the two data sources using the country name of the gapminder dataset and the nation names of the olympic medal tables. However, a challenge of the present country based analysis is how to incorporate the many political changes which happened during the analysis period. As an example, East Germany participated as independent national olympic committee during 1968-1988, but the gapminder data only contain GDP data for Germany as a total. We therefore aggregate the results of the two countries for the analysis. A further important change is the split of the former Soviet Union into several independent states. As a consequence, in 1992 a subset of the former Soviet republics participated as [Unified Team](https://en.wikipedia.org/wiki/Unified_Team_at_the_1992_Summer_Olympics). The GDP values for the Soviet Union thus have to be computed from the Gapminder data by manually summing the individual Soviet republic GDP values. Again we skip further data munging details and simply refer to the GitHub code for a **transparent & reproducible** account. Warning: Only few of the entries in the list of [obsolete nations & name changes ](https://en.wikipedia.org/wiki/All-time_Olympic_Games_medal_table#Notes) are taken into account. 

Conditioned on the success of the previous wrangling step, we can now join the two data sources:

```{r,echo=FALSE}
medals_mod <- medals %>% select(Nation,Year,Gold,Silver, Bronze,Total)

#Merge Germany into one country? Is there a better way to do
#this in dplyr?
germany <- medals_mod %>% group_by(Year) %>% filter(grepl("Germany",Nation)) %>%  summarise_each(funs(sum),-Nation) %>% mutate(Nation="Germany")

medals_mod <- medals_mod %>% filter(!grepl("Germany",Nation)) %>% bind_rows(germany) %>% arrange(Year,desc(Gold))

#Rename Great Britain in the medal dataset to United Kingdom
medals_mod <- medals_mod %>% mutate(Nation = replace(Nation, Nation=="Great Britain", "United Kingdom"))

#Rename the Unified Team in 1992 to Soviet union even though
#countries such as Estonia, Latvia and Lithuania already in 1992
#competed as independent countries
medals_mod <- medals_mod %>% mutate(Nation = replace(Nation, Nation=="Unified Team", "Soviet Union") )

# The gapminder data contain no information on the Soviet Union,
# but the countries making up the former soviet Union are present
# we hence create an artificial Soviet Union GDP based on the
# info in https://en.wikipedia.org/wiki/Soviet_Union
soviet_republics <- c("Russia", "Armenia", "Azerbaijan", 	"Belarus", 	"Estonia", 	"Georgia", 	"Kazakhstan", "Kyrgyzstan", 	"Latvia", "Lithuania", 	"Moldova", 	"Tajikistan", "Turkmenistan", "Ukraine","Uzbekistan")

soviet_union <- gapminder_manual %>% group_by(Year) %>%  filter((Nation %in% soviet_republics & Year < 1992) | (Nation %in% setdiff(soviet_republics, c("Estonia","Latvia","Lithuania")) & Year == 1992)) %>% summarise(GDPpc=mean(GDPpc), Population=sum(Population)) %>% mutate(Nation = "Soviet Union", Continent="FSU")

#Handle Yugoslavia
yugo_republics <- c("Croatia", "Bosnia-Herzegovina", "Slovenia", "Serbia", "Macedonia", "Montenegro", "Bosnia and Herzegovina")

yugoslavia <- gapminder_manual %>% group_by(Year) %>%  filter((Nation %in% yugo_republics & Year < 1992) | (Nation %in% setdiff(yugo_republics, c("Croatia", "Bosnia-Herzegovina", "Slovenia")) & Year == 1992)) %>% summarise(GDPpc=mean(GDPpc), Population=sum(Population)) %>% mutate(Nation = "Yugoslavia", Continent="Europe")

#Update the gapminder dataset
gapminder_manual <- bind_rows(gapminder_manual, soviet_union, yugoslavia)

```

```{r,warning=FALSE}
medals_gm <- left_join(medals_mod, gapminder_manual, by=c("Nation","Year"))
```

# Results

First we analyse the [all-time summer olympic medal table](https://en.wikipedia.org/wiki/All-time_Olympic_Games_medal_table) for the period 1960-2016.

```{r}
medals_alltime <- medals_gm  %>% group_by(Nation)  %>% summarise(Total=sum(Total))  %>% arrange(desc(Total))
DT::datatable(medals_alltime)
```

We now plot of the total number of medals awarded for each summer games in the period of 1960-2016.
```{r,TOTALMEDALSPERGAME}
nTotal <- medals_gm %>% group_by(Year) %>% summarise(TotalOfGames=sum(Total))
ggplot(nTotal, aes(x=Year,y=TotalOfGames)) + geom_line() + ylab("Total number of medals per Summer Games")
```

A distinct increasing trend is observed in the above figure. Hence, in order to make between-country comparisons over time based on the number of medals won, we normalize the medals by the total number of medals awarded during the corresponding games. The result is stored in the column `Frac`.

```{r}
medals_gm <- medals_gm %>% left_join(nTotal, by="Year") %>% mutate(Frac = Total / TotalOfGames)
```

After all these pre-processing steps, we can now compare country results for all summer games in the period 2000-2016.

```{r,FACET2000ANDBEYOND,echo=FALSE,warning=FALSE, fig.width=10,fig.height=6,dpi=200}
#Add city name for better visualization
olympic_city <- data.frame(Year=olympic_years, City=c("Rome","Tokyo", "Mexico City","Munich", "Montreal", "Moscow","Los Angeles", "Seoul", "Barcelona", "Atlanta", "Sydney", "Athens", "Beijing", "London","Rio"))

medals_gm <- left_join(medals_gm, olympic_city, by="Year") %>% mutate("YearCity"=paste0(Year," - ",City))

#Restrict dataset to Year >= 2000
medals_gm2000 <- medals_gm %>% filter(Year >= 2000)

#Plot
p1 <- ggplot(medals_gm2000,
             aes(x=GDPpc, y=Frac, size = Population, colour= Continent)) +
             geom_point() +
    geom_text(data=medals_gm2000, aes(x=GDPpc, y=Frac, label=Nation), vjust=-1,show.legend=FALSE) +
             scale_x_log10() + scale_y_sqrt(labels = scales::percent) +
  xlab("GDP per Capita") + ylab("Fraction of All Medals")

p1 + facet_grid(. ~ YearCity)
```

Note that for better visualization of the many countries with a small number of medals, an $\sqrt{}$-transform of the y-axis is used.

Finally, we can use the `gganimate` package to visualize the dependence of the total number of medals won in the summer games 1960-2016 as a function of GDP per capita. 

```{r,GAPMINDERANIM,fig.width=8,fig.height=5,echo=FALSE,warning=FALSE,results='hide',message=FALSE,fig.keep='hide'}
p2 <- ggplot(medals_gm, aes(x=GDPpc, y=Frac, size = Population, color = Continent, frame = YearCity)) +
  geom_point() +
  geom_text(data=medals_gm, aes(x=GDPpc, y=Frac, label=Nation), vjust=-0.7,show.legend=FALSE, size=1.4*(medals_gm$Population/1e6)^(1/4)+0.2, check_overlap=TRUE) +
  scale_x_log10() +
  xlab("GDP per Capita") + ylab("Fraction of All Medals") +
   scale_y_sqrt(labels = scales::percent)
  
#Animate the result
#gganimate::gg_animate(p2,interval = 3, ani.width=600, ani.height=350)
gganimate::gg_animate(p2, filename="olympicMedals-gapminder-style.gif",interval = 3, ani.width=600, ani.height=350)
```

```{r,results='asis',echo=FALSE}
invisible(file.rename("olympicMedals-gapminder-style.gif",file.path(fullFigPath,"olympicMedals-gapminder-style.gif")))
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"olympicMedals-gapminder-style.gif"),")")
```

As before a $\sqrt{}$-transform of the y-axis is used for better visualization. One interesting observation we see from the animation is that the home-country of the Olympics always appears to do well in the following Olympics. Also note that the [1980](https://en.wikipedia.org/wiki/1980_Summer_Olympics_boycott) and [1984](https://en.wikipedia.org/wiki/1984_Summer_Olympics_boycott) were special due to boycotts. With respect to the top-5 nations it is also worth noticing that China, due to protests against the participation of Taiwan, did not participate in the Olympics 1956-1980. Furthermore, up to 1988 the team denoted "Germany" in the animation consists of the combined number of medals of "East Germany" and "West Germany".

### Fun with Flags

Update: After being made aware of the concurrent [blog entry](http://pmassicotte.github.io/2016-08-25-olympics2016) by [Philippe Massicotte](https://www.researchgate.net/profile/Philippe_Massicotte) on how to visualize the Rio medal table using the `ggflags` package, the above gapminder visualization can easily be extended to use flags instead of nation names. As the `ggflags` package only contains the flags of currently existing countries we start the visualization in 1990. For better visability we also add the trajectory of each nation.

```{r,echo=FALSE,results='hide',warning=FALSE,message=FALSE}
#Use the ggflags package - not on CRAN
#devtools::install_github("baptiste/ggflags")
library("ggflags")
library("countrycode")

medals_gm <- medals_gm %>% mutate(code= countrycode(Nation, "country.name", "iso2c")) %>% mutate(code = tolower(code))
medals_gm %>% select(code)

#Plot starting year 1990 (only then flags & nations match)
medals_gm1990 <- medals_gm %>% filter(Year >= 1990)

#gganimate doesn't work with ggflags, instead create the animation manually
library("animation")

##Safe animation as GIF. Alternative for power point: safeSWF
saveGIF( {
  for (y in olympic_years[olympic_years >= 1990]) {
    #All medals up to now
    medals_up2now <- medals_gm1990 %>% filter(Year <= y)

    p4 <- ggplot(data=medals_up2now, aes(x=GDPpc, y=Frac, group=Nation)) +
      geom_line(aes(alpha=1/(abs(Year-y)/4 + 1)),linejoin="round") +
       geom_flag(data=medals_gm1990 %>% filter(Year==y), aes(x=GDPpc, y=Frac, country=code), size = 10) + 
      scale_x_log10(limits=c(min(medals_gm$GDPpc,na.rm=TRUE),max(medals_gm$GDPpc,na.rm=TRUE))) + 
      scale_y_sqrt(labels = scales::percent, limits=c(0,max(medals_gm1990$Frac))) +
       guides(alpha=FALSE) + 
      xlab("GDP per Capita") + ylab("Fraction of All Medals") + ggtitle(medals_gm1990 %>% filter(Year==y)  %>% select(YearCity)  %>% slice(1) %>% unlist)
    print(p4)
  }
}, movie.name="olympicMedals-flags.gif",ani.width=600, ani.height=350)
```

```{r,results='asis',echo=FALSE}
invisible(file.rename("olympicMedals-flags.gif",file.path(fullFigPath,"olympicMedals-flags.gif")))
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"olympicMedals-flags.gif"),")")
```


### Number of Medals per Population

To see the medal tables in a different light, we instead visualize a quantity relative to the number of medals per population. To enable cross-year comparisons we therefore compute the following index for each country and olympic summer games:
$$
\frac{\text{Fraction of All Medals the Country got in that Year}}{\text{Population in the Country that Year}} \times 10^6.
$$
We shall call this index a country's fraction of all medals per million population. A similar animation as above, now with logarithmic y-axis, illustrates the dynamics. To provide **evidence supported neighbour mocking**, we highlight the position of the three Nordic countries (Denmark, Sweden and Norway).

```{r,GAPMINDERANIMPERPOP,fig.width=8,fig.height=5,echo=FALSE,warning=FALSE,results='hide',message=FALSE,fig.keep='hide'}
#Make a per-population variable
medals_gm <- medals_gm %>% mutate(FracPerPop=Frac/Population * 1e6)
highlight_countries <- c("Denmark","Sweden","Norway")

p3 <- ggplot(medals_gm, aes(x=GDPpc, y=FracPerPop, size = Population, color = Continent, frame = YearCity)) +
  geom_point(data=medals_gm, aes(x=GDPpc, y=FracPerPop, size = Population, color = Continent, frame = YearCity, alpha=ifelse(Nation %in% highlight_countries, 1,0.4)), show.legend=FALSE) +
  geom_text(data=medals_gm, aes(x=GDPpc, y=FracPerPop, label=Nation), vjust=-0.7,show.legend=FALSE,size=ifelse(medals_gm$Nation %in% highlight_countries, 6, 4),alpha=ifelse(medals_gm$Nation %in% highlight_countries, 1,0.4)) +
  scale_x_log10() +
  xlab("GDP per Capita") + ylab("Fraction of All Medals per Million Population") +
   scale_y_log10()

#Animate the result
#gganimate::gg_animate(p3,interval = 3, ani.width=600, ani.height=350)
gganimate::gg_animate(p3, filename="olympicMedals-perpop-gapminder-style.gif",interval = 3, ani.width=600, ani.height=350)
```


```{r,results='asis',echo=FALSE}
invisible(file.rename("olympicMedals-perpop-gapminder-style.gif",file.path(fullFigPath,"olympicMedals-perpop-gapminder-style.gif")))
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"olympicMedals-perpop-gapminder-style.gif"),")")
```

Jamaica, Bahamas and Grenada appear to do reasonably well lately compared to their population size. However, more more important - did you noticed the position of Denmark at the 2016 games in Rio?


