---
layout: post
title: "Cartograms with R"
tags: [dataviz, rstats, spatial, GIS]
bibliography: ~/Literature/Bibtex/jabref.bib
comments: true
---

```{r,include=FALSE,echo=FALSE,message=FALSE}
##If default fig.path, then set it.
if (knitr::opts_chunk$get("fig.path") == "figure/") {
  knitr::opts_knit$set( base.dir = '/Users/hoehle/Sandbox/Blog/')
  knitr::opts_chunk$set(fig.path="figure/source/2016-10-10-cartograms/")
}
fullFigPath <- paste0(knitr::opts_knit$get("base.dir"),knitr::opts_chunk$get("fig.path"))
filePath <- "/Users/hoehle/Sandbox/Blog/figure/source/2016-10-10-cartograms/"

knitr::opts_chunk$set(echo = TRUE,fig.width=8,fig.height=5,fig.cap='')
options(width=90)
library("dplyr")
library("ggplot2")
library("tidyr")
library("methods")
theme_set(theme_bw())
```

## Abstract

We show how to create
[cartograms](https://en.wikipedia.org/wiki/Cartogram) with R by
illustrating the population and age-distribution of the planning
regions of Berlin by static plots and animations.

```{r,results='asis',echo=FALSE}
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"CARTOGRAM-1.png"),")")
```

{% include license.html %}

## Introduction

Every good lecture on sophisticated statistical modelling starts with
underlining the importance of **data visualization** as the first step
of an
analysis. [Choropleth maps](https://en.wikipedia.org/wiki/Choropleth_map)
are a common choice for visualizing the spatial distribution of a
feature recorded in administrative regions, e.g., population density
or the incidence rate of a disease. Here, each region is shaded with a
color selected in accordance with the feature variable, e.g., higher
hue if the feature value is higher. Choosing the right palette for
such visualizations is a science of its own, see
e.g. @zeileis_etal2009 or the [ColorBrewer](http://colorbrewer2.org/)
project, which is available in R through the
[`RColorBrewer`](https://cran.r-project.org/web/packages/RColorBrewer/index.html)
package. A nice way to further spice up your spatial visualizations
are **area cartograms**, where the boundary shape of each region is
warped such that its area becomes proportional to the value of the
feature variable you want to illustrate. The difficult part here is to
preserve the arrangement of the regions, see for example
@gastner_newman2004 for the methodological challenges of this task.

In this post we show how such area cartograms can easily be created
with R using the packages `Rcartogram` and `getcartr` together with
the powerful packages `sp`, `rgeos` and `rgdal` for the spatial data
wrangling. Both `Rcartogram` and `getcartr` are only available from
github, because the license of the underlying
[`Cart`](http://www-personal.umich.edu/~mejn/cart/) C fragment
implementing the method of @gastner_newman2004 does not appear to be
GPL (or the like) compatible.

# The Data

We use population numbers for the 447 planning regions of Berlin
(Lebensweltlich orientierte Räume (LOR)). The boundaries of these
regions are available as ESRI Shapefile through the
[open data portal of Berlin](http://daten.berlin.de/datensaetze/rbs-lor-lebensweltlich-orientierte-r%C3%A4ume-dezember-2015)
under the CC BY license. The 2015 population data of the LORs are
available as CSV file through the same
[data portal](http://daten.berlin.de/datensaetze/einwohnerinnen-und-einwohner-berlin-lor-planungsr%C3%A4umen-am-31122015).

```{r,cache=TRUE}
tmpfile <- paste0(tempfile(),".zip")
download.file("https://www.statistik-berlin-brandenburg.de/opendata/RBS_OD_LOR_2015_12.zip",destfile=tmpfile)
unzip(tmpfile,exdir=file.path(filePath,"RBS_OD_LOR_2015_12"))
download.file("https://www.statistik-berlin-brandenburg.de/opendata/EWR201512E_Matrix.csv",destfile=file.path(filePath,"EWR201512E_Matrix.csv"))
```

With the help of the `rgdal`, `sp` and the `rgeos` CRAN packages, R
can be used as a [geographic information system (GIS)](https://en.wikipedia.org/wiki/Geographic_information_system). This
allows for easy merging of these two data sources together with a
spatial aggregation to the **Prognoseräume** level, which is a
slightly higher level of aggregation than the LORs (60 regions instead
of 447). The output of these data wrangling steps will be a
SpatialPolygonsDataFrame object `pgrs` - see GitHub code for details.

```{r,message=FALSE}
library(rgdal)
library(sp)
library(rgeos)
```
```{r,message=FALSE,warning=FALSE}
##Read shapefile
lor <- readOGR(dsn=file.path(filePath,"RBS_OD_LOR_2015_12"),layer="RBS_OD_LOR_2015_12")
proj4string(lor)
##Compute area of each LOR in km^2 area (unit: meters -> convert to square km)
lor$area <- gArea(lor, byid=TRUE) / (1e6)

##Read population
pop <- readr::read_csv2(file=file.path(filePath, "EWR201512E_Matrix.csv"))

##Merge SpatialPolygonsDataFrame with population information
lor_pop <- merge(lor, pop, by.x="PLR",by.y="RAUMID")
```
````{r,echo=FALSE,results='hide'}
#Make more adequate names for the PGRs -> extended PGR name
lor_pop$EXTPGRNAME <- as.character(lor_pop$PGRNAME)
lor_pop$EXTPGRNAME <- gsub("Region ","Steglitz-Zehlendorf ",lor_pop$EXTPGRNAME)
lor_pop$EXTPGRNAME <- gsub("^SPA ","Spandau ",lor_pop$EXTPGRNAME)
lor_pop$EXTPGRNAME <- gsub("^CW ","Charl.-Wilm. ",lor_pop$EXTPGRNAME)

## Aggregate to level of Prognoseräume (there are 60 of them)
## Population numbers and area is aggregated using sum, extended PGR
## name will be the same in each area and hence one can take the first
aggr_fun <- function(x) {
  if (is.numeric(x)) {
    return(sum(x,na.rm=TRUE))
  } else {
    return(x[1])
  }
}

##Do the aggregation
pgrs <- aggregate(lor_pop[,grep("^E_|^area|^EXTPGRNAME",colnames(as.data.frame(lor_pop)))], by=list(PGR=lor_pop$PGR.x), FUN=aggr_fun)

##Add a population density variable
pgrs$density <- pgrs$E_E / pgrs$area
```

```{r,echo=FALSE}
##check the area computation:
#sum(lor$area) ##should be approx equal to 891,8 km^2 according to https://en.wikipedia.org/wiki/Berlin)
```

Plotting the result of the `pgrs` object as an instance of
`SpatialPolygonsDataFrame` can be done using the standard `Spatial*`
plotting routines documented extensively in, e.g, @bivand_etal2008 and
its comprehensive [webpage](http://www.asdar-book.org/).

```{r,CHOROPLETH,message=FALSE,dpi=200}
######################################################################
## Plotting the result, see nice tutorial by
## http://www.nickeubank.com/wp-content/uploads/2015/10/RGIS3_MakingMaps_part1_mappingVectorData.html
## or the Bivand et al. (2008) book - a must read!
## Note: there is a 2nd edition available nowadays.
######################################################################

library(RColorBrewer)
my.palette <- brewer.pal(n = 6, name = "Purples")
##Helper function for making labels for each entry
sp.label <- function(x, label) {list("sp.text", coordinates(x), label,cex=0.5)}
borderCol <- "white"

#Plot choropleth map
spplot(pgrs, "density", col.regions = my.palette, cuts = length(my.palette)-1, col = borderCol,main="Choropleth map of Population Density",sp.layout=sp.label(pgrs, pgrs$EXTPGRNAME))
require(grid)
grid.text(expression("Population density (Persons / "~km^2~")"), x=unit(0.95, "npc"), y=unit(0.50, "npc"), rot=-90)
```

## Installing the Cartogram R packages

Two packages `Rcartogram` and `getcartr` make the functionality of the
@gastner_newman2004 procedure available for working with
objects of class `Spatial*`. Installing `Rcartogram` requires the
[`fftw` library](http://www.fftw.org/) to be installed. How to best do
that depends on your system, for Mac OS X the
[homebrew package system](http://brew.sh/) makes this installation easy.

```{r,eval=FALSE}
##On command line in OS/X with homebrew. Wrapped in FALSE statement to not run system() unintentionally
if (FALSE) {
  system("brew install fftw")
}
##Install the R implementation of Cart by Gastner and Newman (2004)
devtools::install_github("omegahat/Rcartogram")
devtools::install_github('chrisbrunsdon/getcartr',subdir='getcartr')
```
We are now ready to compute our first cartogram using the
`getcartr::quick.carto` function.

```{r,CARTOGRAM,message=FALSE,dpi=200}
library(Rcartogram)
library(getcartr)

##Make a cartogram
pgrs_carto <- quick.carto(spdf=pgrs,v=pgrs$E_E,res=256)

##Display it using sp functionality
spplot(pgrs_carto, "area", col.regions = my.palette, cuts = length(my.palette)-1, col = borderCol,main="Population Cartogram as Choropleth of Area",sp.layout=sp.label(pgrs_carto, label=pgrs_carto$EXTPGRNAME))
grid::grid.text(expression("Area (km"^2*")"), x=unit(0.95, "npc"), y=unit(0.50, "npc"), rot=-90)
```

```{r,echo=FALSE}
##Find the selection of col names to do the operation on. These are the different age groups
cols <- grep("E_E[0-9]",colnames(as.data.frame(pgrs)),val=TRUE)
cols <- cols[!grepl("U",cols)]
```

With the cartogram functionality now being directly available through
R allows one to embedd cartogram making in a full R pipeline. We
illustrate this by generating a sequence of cartograms into an
animated GIF file using the `animation` package. The animation below
shows a cartogram for the population size for each of the `r length(cols)`
age groups in the Berlin data set. One observes that the
25-45 year old tend to live in the city centre, while the 95-110 year
old seem to concentrate in the wealthy regions in the south west.

```{r,echo=FALSE,cache=TRUE}
##Warp based on each age strata
cartos <- pbapply::pblapply(cols, function(agegrp) {
  ##Zero population maps don't work
  pgrs[[agegrp]] <- ifelse(pgrs[[agegrp]] > 0, pgrs[[agegrp]], 0.1)
  ##Warp
  pgrs_warp <- quick.carto(pgrs,pgrs[[agegrp]])
  return(pgrs_warp)
})
```

```{r,echo=FALSE,results='hide',message=FALSE}
##Animate the warp
library(animation)
ani.options(interval=0.6)

##For some wicked reason writing movie.name =
##file.path(filePath,"pop-cartograms.gif") doesn't work. So we instead
##produce the file in the same directory and move it to the right
##place

theAnimation <- "pop-cartograms.gif"


## Make an animation by looping over all age groups.
saveGIF({
  for (i in 1:length(cartos)) {
    agegrp <- cols[i]
    cat(agegrp,"\n")
    pgrs_carto <- cartos[[i]]
    agename <- paste0("Population age ",gsub("_","-",gsub("E_E","",agegrp)),"y")

    par(mar=c(0, 0, 3, 0) + 0.1)
    plot(pgrs_carto,
         xlim=bbox(cartos[[1]])[1, ], ylim = bbox(cartos[[1]])[2, ],
         main=paste0(agename, " Cartogram"),col="#DADAEB",border="white",lwd=2)
    text(coordinates(pgrs_carto)[,1],coordinates(pgrs_carto)[,2],pgrs$EXTPGRNAME,cex=.45,col="darkgray")
  }
}, movie.name=theAnimation,ani.width=700,ani.height=700,ani.dev=function(...) png(...,res=125))
```


```{r,results='asis',echo=FALSE}
invisible(file.rename(theAnimation,file.path(fullFigPath,theAnimation)))
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),theAnimation),")")
```

# Outlook

While writing this posts some other useRs have posted on how to create
[interactive cartograms](https://twitter.com/Victpir/status/785852075129315333).

# References


