---
layout: post
title: "Happy pbirthday class of 2016"
tags: [rstats, stats, data journalism, onomastics]
bibliography: ~/Literature/Bibtex/jabref.bib
comments: true
---

```{r,include=FALSE,echo=FALSE,message=FALSE}
##If default fig.path, then set it.
if (knitr::opts_chunk$get("fig.path") == "figure/") {
  knitr::opts_knit$set( base.dir = '/Users/hoehle/Sandbox/Blog/')
  knitr::opts_chunk$set(fig.path="figure/source/2017-02-13-bday/")
}
fullFigPath <- paste0(knitr::opts_knit$get("base.dir"),knitr::opts_chunk$get("fig.path"))
filePath <- "/Users/hoehle/Sandbox/Blog/figure/source/2017-02-13-bday/"

knitr::opts_chunk$set(echo = TRUE,fig.width=8,fig.height=4,fig.cap='',fig.align='center',echo=FALSE,dpi=72*2) # autodep=TRUE
options(width=90)
library("dplyr")
library("ggplot2")
library("tidyr")
library("methods")
library("magrittr")

theme_set(theme_bw())
```

```{r,echo=FALSE,results='hide'}
##Load the store results
theFile <- file.path(filePath, "theAnswer.Rdata")
if (file.exists(theFile)) {
  theAnswer <- load(file.path(filePath, "theAnswer.Rdata"))
} else {
  n <- 26L
  p_theAnswer <- 0.34
  pb_up <- list(prob=1, tList=matrix(NA, nrow=10,ncol=10), P=NA,a=NA)
}
```

## Abstract

Continuing the analysis of first names given to newborns in Berlin
2016, we solve the following problem: what is the probability, that in
a school class of size $n$ of these kids there will be at least two kids
having the same first name?  The answer to the problem for classes of
size `r n` is `r sprintf("%.0f%%",100*p_theAnswer)` and can be solved
as an instance
of the birthday problem with unequal probabilities. R code is provided
for solving the problem exactly for moderate $n$ and approximately for
larger $n$. For the case that all probabilities are equal, our results
are compared to the output of R's lovely `pbirthday` function.

<center>
```{r,results='asis',echo=FALSE}
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"APPROXVSEXACT-1.png"),")")
```
</center>

{% include license.html %}

## Introduction

The previous post
[Naming Uncertainty by the Bootstrap](http://staff.math.su.se/hoehle/blog/2017/02/06/onomastics.html)
contained an analysis of first names given to newborns in
Berlin 2016 [@opendataberlinNames2016]. For instance, Marie and Alexander were
the top names among girls and boys, respectively. In a comment
[Maëlle](http://www.masalmon.eu/) asked *what's the resulting
probability that there will be kids with the same first name in a school
class*? We implement equations by @klotz1979 and @mase1992 in R in
order to answer this important question.

## The Birthday Problem

The above posed question is a variation of the
[**birthday problem**](https://en.wikipedia.org/wiki/Birthday_problem),
which every statistician has solved at least once in an introductory
probability class: *in a class of $n$ randomly chosen persons, what is
the probability that some pair of them will have the same birthday*?
Assuming that there are $N=365$ possible birthdays and all birthdays are
equally probable the answer can be calculated as:

$$
P(\text{at least two people in the class have the same birthday}) =
1-\frac{(N)_{n}}{N^n},
$$

where $(x)_n = x! / (x-n)!$ is the so called **factorial polynomial**. Say we are
interested in $n=26$, which is the
[maximal allowed class size](https://www.berlin.de/imperia/md/content/sen-bildung/rechtsvorschriften/grundschulverordnung.pdf)
in Berlin's elementary schools (§4, Sect. 8 in the regulation). We can perform the
necessary calculations either directly or by R's `pbirthday`function.

```{r, echo=TRUE}
n <- 26 ; N <- 365
c(manual=1 - exp(lfactorial(N)-lfactorial(N-n) - n*log(N)),
  pbirthday=pbirthday(n=n,classes=N))
```

Finding the `pbirthday` function as part of base R was a bit
surprising, but just underlines that R really has its roots in
**statistics**!

### Birthday Problem with Unequal Probabilities

```{r, READDATA, message=FALSE}
##Baby names in Berlin
filePath_onomastics <- "/Users/hoehle/Sandbox/Blog/figure/source/2017-02-06-onomastics/"
library(readr)
year <- 2016
distrNames <- read_csv(file=file.path(filePath_onomastics,paste0("berlin-firstnames-",year,".csv"))) %>% select(-strata)
distrNames <- distrNames %>% mutate(district = factor(district), sex=factor(sex))
```

In our problem $N$ corresponds to all possible names of newborns
in 2016. For the analysis we only group by first name and thus do not
distinguish between instances of the same name used for both
girls and boys.

```{r,echo=TRUE}
newborn <- distrNames %>% group_by(firstname) %>%
  summarise(count=sum(count)) %>% ungroup() %>%
  mutate(p=count/sum(count)) %>%
  arrange(desc(count))
```
```{r,echo=FALSE}
print(newborn, n=4)
```

In total there are $N=`r nrow(newborn)`$ possible names. From the $p$
column it also becomes obvious that not all names are equally
likely. Had they been, the quick solution to Maëlle's question would
have been:

```{r,echo=TRUE}
pbirthday(n=26, classes=nrow(newborn))
```

```{r,echo=FALSE, results='hide'}
(N <- nrow(newborn))
```
Less than `r sprintf("%0.f%%",ceiling(100*pbirthday(n=26, classes=nrow(newborn))))`!
However, we expect this probability to be much higher, if we start to
take the unequal occurrence probabilities into account. So let's do it!

It's easy to see that the probability of no collision, i.e. no kids
having the same name in the class, can be calculated as:
$$
P(\overline{C}_n) = n!
\underset{1\leq i_1 < i_2 < \cdots <i_n \leq N}{\sum \sum \cdots \sum} \>
p_{i_1} p_{i_2} \cdots p_{i_n}.
$$

However, this is a formidable number of terms to sum. In the
case of $N=`r N`$ and $n=`r n`$ the number is:

```{r,eval=FALSE}
##noquote(sprintf("%-50.0f", choose(N,n)))
```
```{r,echo=TRUE}
Rmpfr::chooseMpfr(N,n)
```
That's an `r ceiling(log10(choose(N,n)))` digit number! This is
not *ever* going to happen. Instead @klotz1979, based on generating
functions, showed that the above equation corresponds to

$$
P(\overline{C}_n) = n!
\underset{\underset{\sum_{j=1}^n j \cdot t_j = n}{0\leq t_1,t_2,\ldots,t_n \leq n}}{\sum \sum \cdots \sum}
(-1)^{n + \sum_j t_j}
\left(
\prod_{j=1}^n \frac{ (P_j/j)^{t_j}}{t_j!}
\right),
$$
where $P_j = \sum_{i=1}^N p_i^j$. Let the vector
$\mathbf{t}=(t_1,\ldots,t_n)$ count the number of singletons ($t_1$),
doubletons ($t_2$), triplets ($t_3$), ..., up to the number of names
occurring $n$ times ($t_n$). The above sum means that we have to sum
over all $\mathbf{t}$ such that $\sum_{j=1}^n j \cdot t_j = n$. The
number of such terms to sum is much lower than in the previous
expression, e.g., for $N=`r nrow(newborn)`$
and $n=`r n`$ the number of terms is `r nrow(pb_up$tList)`.


The above computations have been made available in the R package
[`birthdayproblem`]((https://github.com/hoehleatsu/birthdayproblem))
available from github:

```{r,echo=TRUE,results="hide",warning=FALSE,message=FALSE}
devtools::install_github("hoehleatsu/birthdayproblem")
```

As an example, for $n=4$ all the necessary terms to sum can be found
somewhat brute-force'ish by running through the following four nested
for loops:
```{r}
##No so pretty
cat(paste0(paste0(birthdayproblem:::make_tListFunc_syntax(n=4),collapse="\n"),"\n"))
```
This function would then return the necessary sets for the $n=4$ case:
```{r}
source(textConnection(birthdayproblem:::make_tListFunc_syntax(n=4)))
compute_tList()
```
which can be processed further as in the @klotz1979 equation stated
above in order to compute the probability of interest.

In the accompanying
[R code of this blog post](https://github.com/hoehleatsu/hoehleatsu.github.io/blob/master/_source/2017-02-13-bday.Rmd)
the above $n$ nested loops
are constructed by the function `birthdayproblem:::make_tListFunc_syntax`, which given
$n$ generates the syntax of the necessary function nested loop
function. Calling `source` on this syntax string then provides a
proper R function to evaluate. A similar function
`birthdayproblem:::make_tListFunc_syntax_cpp` is provided to generate an equivalent C++
function, which then using Rcpp's `sourceCpp` function can be turned
into an R function. As a **side note**: The nested for loops for
increasing $n$ quickly look foul, which earned it the predicate
*possibly the best nested loop ever* in a comment of a
[stackoverflow post](http://stackoverflow.com/questions/42130954/compiling-many-nested-loops-using-the-rcppcppfunction)
concerned with the many nested loops breaking the `clang` compiler on
MacOS.

The above described syntax generation, evaluation and post-processing
steps necessary to compute the desired probability
$1-P(\overline{C}_n)$ are all
implemented in the function `birthdayproblem::pbirthday_up` (postfix: `up` for
*u*nequal *p*ropabilities) in honour of R's `pbirthday` function.  A `method`
argument allows the user to choose if the nested-loops should be
computed using `"R"`, `"Rcpp"`. As an alternative to the this exact
solution by @klotz1979 one can also compute an approximate solution
described in @mase1992, which is of the impressive order
$O(1)$ while being extremely accurate (use `method="mase1992"`).
The R method works in
acceptable time for $n$'s up to around 35, the Rcpp runs $n=60$ in
less than three minutes; for larger $n$ the approximation is to be
recommended if you don't like waiting.

With all code in place we finally can provide Maëlle with the correct
answer to her question:
```{r,echo=TRUE}
n <- 26L
(p_theAnswer <- birthdayproblem::pbirthday_up(n=n, prob=newborn %$% p)$prob)
```
```{r}
pb_up <- birthdayproblem::pbirthday_up(n=n, prob=newborn %$% p)
save(file=file.path(filePath,"theAnswer.RData"),list=c("n","p_theAnswer","pb_up"))
```
In other words, the probability of having a name collision in a class
of $n=`r n`$ is `r sprintf("%.1f%%",p_theAnswer*100)`. If local politics
would decide to increase the maximum class size by one, the resulting
probability for $n=27$
would be:
`r sprintf("%.1f%%",100*birthdayproblem::pbirthday_up(n=n+1L, prob=newborn %$% p)$prob)`.
One more reason against increasing [school class size](http://www.nzherald.co.nz/nz/news/article.cfm?c_id=1&objectid=11288618)?


### Numerical Comparisons

We first test the `birthdayproblem` package's `pbirthday_up` function on the classical birthday
problem with equal probabilities:
```{r,echo=TRUE, warning=FALSE, message=FALSE}
c(pbirthday=pbirthday(n=26L, 365),
  klotz1979_R=birthdayproblem::pbirthday_up(n=26L, rep(1/365, 365), method="R")$prob,
  klotz1979_Rcpp=birthdayproblem::pbirthday_up(n=26L, rep(1/365, 365), method="Rcpp")$prob,
  mase1992=birthdayproblem::pbirthday_up(n=26L, rep(1/365, 365), method="mase1992")$prob)
```
works like a **dream**!

```{r, warning=FALSE, message=FALSE, results='hide', cache=TRUE}
n <- 40L
##Speed compare
timeR <- system.time(print(birthdayproblem::pbirthday_up(n=n, prob=newborn %$% p)$prob))
timeRcpp <- system.time(print(birthdayproblem::pbirthday_up(n=n, prob=newborn %$% p, method="Rcpp")$prob))
timeMase1992 <- system.time(print(birthdayproblem::pbirthday_up(n=n, prob=newborn %$% p, method="mase1992")$prob))
```

Speed-wise, the R looping approach takes `r sprintf("%0.f",timeR[3])`s to compute the
result for $n=`r n`$. The Rcpp approach on the other hand works in just
`r sprintf("%.0f",timeRcpp[3])`s. The approximation by @mase1992 only takes
`r sprintf("%.3f",timeMase1992[3])` s. To assess the quality of the
approximation we consider a range of different $n$:

```{r,cache=TRUE, warning=FALSE, message=FALSE,results='hide'}
birthday <- data.frame(n=c(1,5,seq(10,50,by=5)))
birthday <- birthday %>% rowwise %>% do( {
  n <- as.integer(.$n)
  p <- newborn %$% p
  p_k <- birthdayproblem::pbirthday_up(n=n, prob=p, method=ifelse(n<=35,"R","Rcpp"))$prob
  p_m <- birthdayproblem::pbirthday_up(n=n, prob=p, method="mase1992")$prob
  data.frame(n=n,p=c(p_k,p_m), abs=abs(p_m-p_k), method=c("exact_klotz1979","approx_mase1992"))
})
```

```{r, APPROXVSEXACT, warning=FALSE, message=FALSE}
##Show the result as a function of n
ggplot( birthday, aes(x=n, y=p, color=method)) + geom_line(size=1.3) + ylab("Probability") + ggtitle(paste0("Probability of at least two kids with the same first name in a class of n")) + scale_y_continuous(labels=scales::percent) + scale_color_discrete(name="Method")
```

It's hardly possible to see the difference between the approximation
and the exact solution. For better comparison, we also show the
absolute error between the approximate solution and the exact
solution:

```{r, ABSERROR, warning=FALSE, message=FALSE}
b2 <- birthday %>% filter(n>1) %>% group_by(n) %>% summarise(abs=max(abs))
ggplot( b2, aes(x=n, y=abs)) + geom_line() + ylab("Absolute error")
```

It's amazing to see how small the error really is.

## Discussion

```{r,results='hide'}
theAnswer <- load(file.path(filePath, "theAnswer.Rdata"))
```
We calculated that the probability of a name-collision in a class of
$n=`r n`$ kids born in Berlin 2016 is
`r sprintf("%.0f%%",100*p_theAnswer)`. Furthermore, we showed that clever
mathematical approximations are better than brute-force computations, that
stack-exchange rules and that Rcpp can speed up your R program
considerably. Furthermore, you have been shown the best nested for loop
ever! Finally, in honour of
[Jerome Klotz](https://www.secfac.wisc.edu/senate/2007/0305/1976(mem_res).pdf)
a screenshot of the
Acknowledgements section of the
@klotz1979 technical report:

<center>
```{r,results='asis',echo=FALSE}
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"klotz1979.png"),")")
```
</center>
<p>
<p>

## Literature
