---
layout: post
title: "On a First Name Basis with Statistics Sweden"
tags: [rstats, stats, data journalism, onomastics]
bibliography: ~/Literature/Bibtex/jabref.bib
comments: true
---


```{r,include=FALSE,echo=FALSE,message=FALSE}
##If default fig.path, then set it.
if (knitr::opts_chunk$get("fig.path") == "figure/") {
  knitr::opts_knit$set( base.dir = '/Users/hoehle/Sandbox/Blog/')
  knitr::opts_chunk$set(fig.path="figure/source/2017-03-25-scbnames/")
}
fullFigPath <- paste0(knitr::opts_knit$get("base.dir"),knitr::opts_chunk$get("fig.path"))
filePath <- file.path("/","Users","hoehle","Sandbox", "Blog", "figure", "source", "2017-03-25-scbnames")

knitr::opts_chunk$set(echo = TRUE,fig.width=8,fig.height=4,fig.cap='',fig.align='center',echo=FALSE,dpi=72*2) # autodep=TRUE
options(width=90)

library(dplyr)
library(ggplot2)
library(magrittr)

theme_set(theme_bw())
```

## Abstract

Judging on the amount of blog posts, many data scientists seem to be
concerned with scraping data from various media sources (Wikipedia,
twitter). However, one should be aware that well structured and high
quality dataset are available through state and country statistics
offices. Increasingly these are offered to the public through direct
database access, e.g., using a REST API. We illustrate the usefulness
of such an approach by accessing data from Statistics Sweden.

<center>
```{r,results='asis',echo=FALSE}
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"COLLISION26.png"),")")
```
</center>

{% include license.html %}

## Introduction

Scandinavian countries are well known to be world-class when it comes
to public registries. So when in need for reliable population data,
this is the place to look. As an example, we access Statistics Sweden
data by API using the `pxweb` package developed by
[\@MansMeg](https://twitter.com/MansMeg),
[\@antagomir](https://twitter.com/antagomir) and
[\@LCHansson](https://twitter.com/LCHansson).  Love was the first
speaker at a Stockholm R-Meetup some
[years ago](https://www.meetup.com/StockholmR/events/105738342/),
where I also gave a talk. Funny how such R-Meetups become useful many years
after!

```{r,echo=TRUE, message=FALSE,results='hide'}
library(pxweb)
```
```{r}
##pcweb pkg recommends setting the console into UTF-8 encoding
##Sys.setlocale(locale="UTF-8")
```

By browsing the Statistics Sweden (in Swedish: Statistiska
Centralbyrån (SCB)) data using their
[web interface](http://www.scb.se/en/finding-statistics/)
one sees that they have two relevant first name datasets: one
containing the *tilltalsnamn* of newborns udinrg 1998-2016 and one for
the years 2004-2016. **Note**: A
[*tilltalsnamn*](https://translate.google.com/translate?sl=auto&tl=en&js=y&prev=_t&hl=en&ie=UTF-8&u=https%3A%2F%2Fsv.wikipedia.org%2Fwiki%2FTilltalsnamn&edit-text=&act=url)
in Sweden is *the* first name (of several possible first names) by
which a person is usually addressed. About 2/3 of the persons in the
Swedish name registry indicate which of their first names is their
tilltalsnamn. For the remaining persons the tilltalsnamn is
automatically chosen to be
[their initial first name](http://www.scb.se/hitta-statistik/statistik-efter-amne/befolkning/amnesovergripande-statistik/namnstatistik/produktrelaterat/Fordjupad-information/fragor-och-svar-om-namnstatistiken/#fem).
**Also note:** For reasons of data protection, the 1998-2016 dataset
contains only first names used 10 or more times in a given year, the
2004-2016 dataset contains only first names used 2 or more times in a
given year.


Downloading such data through the SCB web-interface is cumbersome,
because the downloads are limited to 50,000 data cells per
query. Hence one has to do several manual queries to get hold of the
relevant data. This is where the
[API](http://www.scb.se/en_/About-us/Open-data-API/API-for-the-Statistical-Database-/)
becomes a real time-saver. Instead of trying to fiddle with the API
directly using `rjson` or `RJSONIO` we use the specially designed
`pxweb` package.  One can either use the web-interface to determine
the name of the desired data matrix to query or navigate directly
through the api using `pxweb`:
```{r,echo=TRUE, eval=FALSE}
d <- interactive_pxweb(api = "api.scb.se", version = "v1", lang = "en")
```

and select *Population* followed by *Name statistics* and then
`BE0001T04Ar` or `BE0001T04BAr`, respectively, in order to obtain the
relevant data or api download url. This leads to the following code:

```{r,cache=TRUE,echo=TRUE, message=FALSE}
names10 <- get_pxweb_data(
  url = "http://api.scb.se/OV0104/v1/doris/en/ssd/BE/BE0001/BE0001T04Ar",
  dims = list(Tilltalsnamn = c('*'),
              ContentsCode = c('BE0001AH'),
              Tid = c('*')),
  clean = TRUE) %>% as.tbl
```

For better usability we rename the columns a little and
replace `NA` counts to zero.

```{r,echo=TRUE, message=FALSE}
##Data dancing
names10 <- names10 %>% select(-observations) %>%
  rename(firstname=`first name normally used`,counts=values) %>%
  mutate(counts = ifelse(is.na(counts),0,counts))
##Look at 10 random lines
names10 %>% slice(sample(seq_len(nrow(names10)),size=5))
```

Note: Each spelling variant of a name in the data is treated as a
unique name. In similar fashion we download the `BE0001AL` dataset as `names2`.
```{r,cache=TRUE,message=FALSE}
names2 <- get_pxweb_data(
  url = "http://api.scb.se/OV0104/v1/doris/en/ssd/BE/BE0001/BE0001T04BAr",
  dims = list(Tilltalsnamn = c('*'),
              ContentsCode = c('BE0001AL'),
              Tid = c('*')),
  clean = TRUE) %>% as.tbl %>%
  select(-observations) %>%
  rename(firstname=`first name normally used`,counts=values) %>%
  mutate(counts = ifelse(is.na(counts),0,counts))
```
We now join the two datasets into one large `data.frame`:
```{r,cache=TRUE,echo=TRUE, message=FALSE}
names <- rbind(data.frame(names2,type="min02"), data.frame(names10,type="min10"))
```
and thus got everything in place to compute the name collision
probability over time using the `birthdayproblem` package (as shown in
[previous posts](http://staff.math.su.se/hoehle/blog/2017/02/13/bday.html)).

```{r,echo=TRUE, message=FALSE}
library(birthdayproblem)
collision <- names %>% group_by(year,type) %>% do({
  data.frame(p=pbirthday_up(n=26L, prob= .$counts / sum(.$counts),method="mase1992")$prob, gini= ineq::Gini(.$counts))
}) %>% ungroup %>% mutate(year=as.numeric(as.character(year)))
```

And the resulting probabilities based on the two datasets (`min02`=at
least two instances of the name in a given year, `min10`=at least ten
instances of the name in a given year) can easily be visualized over time.

```{r,COLLISION26,echo=TRUE, message=FALSE}
ggplot( collision, aes(x=year, y=p, color=type)) + geom_line(size=1.5) +
  scale_y_continuous(label=scales::percent,limits=c(0,1)) +
  xlab("Year") + ylab("Probability") +
  ggtitle("Probability of a name collision in a class of 26 kids born in year YYYY") +
  scale_colour_discrete(name = "Dataset")
```

The true probabilities, i.e. taking all tilltalsnamn into account, would be
based on the hypothetical `min1` data set. These probabilities would
be slightly, but not substantially, below the `min2` line. However,
since not every newborn has their *tilltalsnamn* specified, it's not
possible to deduce the total number of kids with a tilltalsnamn
occurring exactly once in a year.  The same problem occurs, e.g., in
the corresponding
[UK and Wales data](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/livebirths/datasets/babynamesenglandandwalesbabynamesstatisticsboys). Here, Table 6 is listing all first names with 3 or more uses, but
not stating how many newborns have a name occurring once and twice,
respectively.

With all due respect for the need to anonymise the name statistics,
it's hard to understand why this summary figure is not automatically
reported, so one would be able to at least compute correct totals or
collision probabilities.  Altogether, I was quite happy to get proper
individual name data so the collision probabilities are - opposite to
some of my [previous blog analyses](http://staff.math.su.se/hoehle/blog/2017/02/13/bday.html) - exact!

```{r,eval=FALSE}
ggplot( collision, aes(x=year, y=gini, color=type)) + geom_line(size=1.5) + scale_y_continuous(label=scales::percent,limits=c(0,1)) + xlab("Year") + ylab("Gini index")
```
