---
layout: post
title: "Confidence Intervals without Your Collaborator's Tears"
tags: [rstats, stats, regulation, EPA]
bibliography: ~/Literature/Bibtex/jabref.bib
comments: true
---


```{r,include=FALSE,echo=FALSE,message=FALSE}
##If default fig.path, then set it.
if (knitr::opts_chunk$get("fig.path") == "figure/") {
  knitr::opts_knit$set( base.dir = '/Users/hoehle/Sandbox/Blog/')
  knitr::opts_chunk$set(fig.path="figure/source/2017-06-22-interpretcis/")
}
fullFigPath <- paste0(knitr::opts_knit$get("base.dir"),knitr::opts_chunk$get("fig.path"))
filePath <- file.path("/","Users","hoehle","Sandbox", "Blog", "figure", "source", "2017-06-22-interpretcis")

knitr::opts_chunk$set(echo = TRUE,fig.width=8,fig.height=4,fig.cap='',fig.align='center',echo=FALSE,dpi=72*2) # autodep=TRUE
options(width=90)

suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(gridExtra))
theme_set(theme_bw())
```

```{r,}
set.seed(42)
n <- 420
x <- rbinom(n=1, size=n, prob=0.1)
```

## Abstract

We provide an interpretation for the confidence interval for a
binomial proportion hidden as the transcript of an hypothetical
statistical consulting session.

<center>
```{r,results='asis',echo=FALSE}
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"BETABINOMIALCI-1.png"),")")
```
</center>

{% include license.html %}

<p>
## The Statistical Consulting Session

**[Client]:** So I did this important experiment where I investigated `r n`
turtles whether they had a specific phenotype trait. `r x` out of my `r n`
turtles had the trait, i.e. the proportion in the sample was `r sprintf("%.1f%%",100*x/n)`.
Now I'd like to state a 95% confidence
interval for what the proportion of the trait is in the population of
turtles (which is pretty large).

**[Statistician]:** Ok, that should be possible. Since your population
is large, we do not need a finite population correction, Since your
sample is also pretty large, let's therefore just go with the textbook
large-sample confidence interval for a binomial proportion, that is $\hat{p}
\pm 1.96 \sqrt{\hat{p}(1-\hat{p})/n}$. Don't worry about the equation,
let's just use R and the `binom` package to compute the interval:

```{r, echo=TRUE}
(ci <- binom::binom.asymp(x=x, n=n, conf.level=0.95))
```

So the requested 95% CI is `r sprintf("%.1f%%",ci["lower"]*100)`-
`r sprintf("%.1f%%",ci["upper"]*100)`. Happy?

**[Client]:** Well, yes, that looks nice, but I wonder how I can
interpret the confidence interval. Is it correct to say that the interval
denotes a region, which contains the true value with 95%
probability?

**[Statistician]:** A confidence interval constructing procedure
yields a random interval, because it depends on quantities (in
particular $x$), which are random. However, once you plug-in the
realization of the random variable (i.e. your observation), you get a
realization of the confidence interval consisting of two numbers.
Would we know the true value then we could tell, whether the value is
covered by this particular interval or not. The specific interval we
calculated above thus contains the true value with probability 0% or
100% - which of the two is the cases we unfortunately do not know.

The correct interpretation is thus that the confidence interval is
constructed by a procedure, which, when you repeat the experiment many
many times, is such that in 95% of the experiments the corresponding confidence
interval would cover the true value. I can illustrate this using a
small simulation study in R. Let's assume your true proportion is 10%.

```{r,echo=TRUE}
##Assumed true value of the proportion of turtles with the trait
theta_true <- 0.1

##Prepare data.frame with 10000 realizations of the experiment
df <- data.frame(x=rbinom(n=1e4, size=n, prob=theta_true))

##Compute 95% CI for each experiment
df %<>% do( {
  as.data.frame(cbind(.,rate=.$x/n,binom::binom.asymp(x=.$x, n=n)[c("lower","upper")]))
})
##Check if the interval covers the true value
df %<>% mutate(covers_true_value = lower < theta_true & upper > theta_true)

##Proportion of intervals covering the true value. This would be 95%
df %>% summarise(coverage = mean(covers_true_value))
```

A graphic for the first 50 experiments shows their point estimates,
corresponding intervals and how they overlap the true value or not:
```{r}
df2 <- df %>% slice(1:20) %>% select(-covers_true_value) %>% gather(type, value, -x)

df20 <- df %>% slice(1:50)
ggplot(df20,aes(x=1:nrow(df20), y=rate)) + geom_point() + geom_linerange(data=df20, ymin=df20$lower, ymax=df20$upper) + ylim(c(0,0.2)) + geom_hline(yintercept=0.1, lty=2, col="indianred2") + xlab("Experiment Number") + ylab("Proportion and 95% CI")
```

The specific confidence interval `ci` we computed above is thus just
one of many possible confidence intervals originating from above procedure.

**[Client]:** Ok, I get that is what happens when you do it many
times. But I just have this one experiment with x = `r x` out of
n = `r n` subjects having the trait of interest. Your above output does
contain a very specific CI, with some very specific
numbers, i.e. `r sprintf("%.1f%%",ci["lower"]*100)`- `r sprintf("%.1f%%",ci["upper"]*100)`.
Is the true value in this interval with 95% probability?

**[Statistician]:** No, the realized interval either covers the true
value or not. But since we do not know the true value it's a bit like
Schrödinger's cat... To keep things very sketchy: The width of the
interval gives you an indication of your estimation certainty, but the
particular values are hard to interpret - except maybe as the critical
limits in a corresponding hypothesis test.

**[Client]:** I'm sorry, but this is **useless**! You suggest to follow a
common statistical protocol, you compute some numbers, but these numbers don't
really mean anything?

**[Statistician]:** Sorry, that's the way it is. However, allow me to
offer a different explanation: We both appear to agree that there is a
true value of the proportion, right? Let's denote this $\theta$ with
$0 \leq \theta \leq 1$.  We don't know the true value, but we might
have some prior idea about the range of plausible values for it. Would
you be willing to characterize your belief about this by a
distribution for $\theta$?  This could be something as simple as just
assuming $\theta \sim U(0,1)$, i.e. you would assume that every value
of $\theta$ between zero and one is equally probable initially. Then
the interval we computed above denotes a 95% equal tail probability
**credibility region** in a Bayesian framework when we assume this flat
prior and use an asymptotic argument resulting in the posterior being
Gaussian with the same mean and variance as the one would get from the
asymptotic frequentist procedure.

**[Client]:** Umm, ok, but how's that helpful?

**[Statistician]:** In the Bayesian context a 95% credibility region
is a summary of your posterior belief about the parameter. Hence, it
is ok to interpret this interval as that your belief after seeing the
data is such that *the true
value is in that interval with 95% probability*.

**[Client]:** I love this Bayes thing! This is what I wanted
initially. But hang on, I don't really believe in all values of the
parameter being initially equally probable. I've seen some
previous studies in the literature who under pretty similar
conditions find that proportion is around 5-15%. How would I
incorporate that?

**[Statistician]:** You could modify the large-sample Gaussian
posterior such that it includes this prior information. Instead of
using Gaussians you could alternatively also use a
[Beta distribution](https://en.wikipedia.org/wiki/Beta_distribution)
to perform so called **conjugate prior-posterior updating** of your
belief about the true proportion. Let's say the 5-15% denotes your
prior 95% credibility region for the parameter. We would use this to
find the parameters of a beta distribution. Then we would update your
prior belief with the observed data to get the posterior
distribution. A feature of such a conjugate approach is that this
updated distribution, i.e. the posterior, is again beta. Doing this in
R is easy:

```{r, echo=TRUE, warning=FALSE}
##Function to determine beta parameters s.t. the 2.5% and 97.5% quantile match the specified values
target <- function(theta, prior_interval, alpha=0.05) {
  sum( (qbeta(c(alpha/2, 1-alpha/2), theta[1], theta[2]) - prior_interval)^2)
}

##Find the prior parameters
prior_params <- optim(c(10,10),target, prior_interval=c(0.05, 0.15))$par
prior_params

```

Actually, you can interpret the above prior parameters as the number of
turtles you have seen with the trait (`r round(prior_params[1])`) and without
the trait (`r round(prior_params[2])`), respectively, before conducting your
above investigation. The posterior parameters are then just
`r round(prior_params[1])` + `r x` = `r round(prior_params[1]) + x` turtles
with the trait and `r round(prior_params[2])` + `r n-x` = `r round(prior_params[2])+(n-x)` turtles without
the trait. You get the credible interval based on this posterior in R by:

```{r, echo=TRUE}
##Compute credibile interval from a beta-binomial conjugate prior-posterior approach
(ci_bayes <- binom::binom.bayes(x=x, n=n, type="central", prior.shape1=prior_params[1], prior.shape2=prior_params[2]))
```

Graphically, your belief updating looks as follows:
```{r, BETABINOMIALCI, echo=TRUE, warning=FALSE, message=FALSE}
##Plot of the beta-posterior
p2 <- binom::binom.bayes.densityplot(ci_bayes)
##Add plot of the beta-prior
df <- data.frame(x=seq(0,1,length=1000)) %>% mutate(pdf=dbeta(x, prior_params[1], prior_params[2]))
p2 + geom_line(data=df, aes(x=x, y=pdf), col="darkgray",lty=2) +
  coord_cartesian(xlim=c(0,0.25)) + scale_x_continuous(labels=scales::percent)
```

The dashed line in the graphic shows the density of your beta prior
distribution. The shaded region shows the density of your beta posterior
distribution, in particular, the gray shaded area denotes your 95%
credibility region on the x-axis.  One advantage of the beta-binomial
approach is that this ensures that the prior as well as the posterior
distribution have the same support as your parameter, i.e. the interval
between zero and one.

**[Client]:** So I can really write that the interval
`r sprintf("%.1f%%",ci_bayes["lower"]*100)`-
`r sprintf("%.1f%%",ci_bayes["upper"]*100)` contains the true parameter with
a probability of 95%?

**[Statistician]:** Depending on the technical level of your readership you might want to state that it's a 95% equi-tailed credible interval resulting from a beta-binomial
  conjugate Bayesian approach obtained when using a prior beta with parameters
  such that the similar 95% equi-tailed **prior credible interval** has limits
  0.05 and 0.15. Given these assumptions the interval
`r sprintf("%.1f%%",ci_bayes["lower"]*100)`-
`r sprintf("%.1f%%",ci_bayes["upper"]*100)`
contains 95% of your subjective posterior density for the parameter.

**[Client]:** ???

**[Statistician]:** You might just want to skip all the details and
  just write that the **95% Bayesian confidence interval** is `r sprintf("%.1f%%",ci_bayes["lower"]*100)`-
`r sprintf("%.1f%%",ci_bayes["upper"]*100)`.

**[Client]:** And what if the reviewers do not agree with my prior distribution?

**[Statistician]:** Then return to your flat prior, i.e. the
uniform. Nobody usually questions that even though you just told me
that you don't believe in it yourself:
```{r, echo=TRUE}
(ci_bayes_unif <- binom::binom.bayes(x=x, n=n, type="central", prior.shape1=1, prior.shape2=1))
```

**[Client]:** Alright, that's settled then - I just use a Bayesian
confidence interval with a flat prior and am done. That was not so
hard after all. Now to something else, the reviewers of the paper,
which by the way has to be revised by tomorrow, also asked us to
substantiate our findings with a **p-value**. How would I get a p-value
related to the above?

**[Statistician]:** Look how late it already is! I really need to
  go. I have this Bayesian Methods class to teach...


