---
layout: post
title: "Amazon's Hanging Cable Problem (Golden Gate Edition)"
tags: [rstats, stats, economics, data visualization, income, equality]
#  bibliography: ~/Literature/Bibtex/jabref.bib
header-includes:
   - \usepackage{bm}
comments: true
editor_options:
  chunk_output_type: console
---

```{r,include=FALSE,echo=FALSE,message=FALSE}
##If default fig.path, then set it.
if (knitr::opts_chunk$get("fig.path") == "figure/") {
  knitr::opts_knit$set( base.dir = '/Users/hoehle/Sandbox/Blog/')
  knitr::opts_chunk$set(fig.path="figure/source/2018-07-23-cable/")
}
fullFigPath <- paste0(knitr::opts_knit$get("base.dir"),knitr::opts_chunk$get("fig.path"))
filePath <- file.path("","Users","hoehle","Sandbox", "Blog", "figure", "source", "2018-07-23-cable")

knitr::opts_chunk$set(echo = TRUE,fig.width=8,fig.height=4,fig.cap='',fig.align='center',echo=FALSE,dpi=72*2) # autodep=TRUE
options(width=150)

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(knitr))

## Packages used for this post
suppressPackageStartupMessages(require(nleqslv))
suppressPackageStartupMessages(require(viridis))

##Configuration
options(knitr.table.format = "html")
theme_set(theme_minimal())
#if there are more than n rows in the tibble, print only the first m rows.
options(tibble.print_max = 10, tibble.print_min = 5)
```

## Abstract:

In this post we use R's capabilities to solve nonlinear equation
systems in order to answer an extension of the hanging cable problem
to suspension bridges. We then use R and ggplot to overlay the
solution to an image of the Golden Gate Bridge in order to bring
together theory and practice.

<center>
```{r,results='asis',echo=FALSE,fig.cap="The cable problem"}
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"GOLDENGATE-1.png"),")")
```
</center>



{% include license.html %}


## Introduction

The so called
[Amazon's hanging cable problem](https://mindyourdecisions.com/blog/2018/07/12/can-you-solve-amazons-hanging-cable-interview-question/) explained in this [youtube video](https://youtu.be/l_ffdarcJiQ)
(watched 2.4 mio times!^[As of 2018-07-23.]) goes as follows:

*A cable of 80 meters (m) is hanging from the top of two poles that are
both 50 m from the ground. What is the distance between the two poles,
to one decimal place, if the center of the cable is*:

(a) *20 m above the ground?*<br>
(b) *10 m above the ground?*<br>

Allegedly, (b) has been used as an Amazon interview question, however, the
problem is much older and has otherwise nothing to do with Amazon. Can you
solve (b)? Or even (a)? The problem can be illustrated as follows:
<p>
<center>
```{r,results='asis',echo=FALSE,fig.cap="The cable problem"}
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"cableproblem.png"),")")
```
</center>
<FONT COLOR="bbbbbb">Screenshot from [Presh Talwalkar's](https://mindyourdecisions.com/blog/2018/07/12/can-you-solve-amazons-hanging-cable-interview-question/) website.</FONT>
<p>

Hint: The
[solution to (a)](https://mindyourdecisions.com/blog/2018/07/12/can-you-solve-amazons-hanging-cable-interview-question/)
is concisely described in @chatterjee_nita2010 and for (b) you need to
do little more than just think. So instead of applying at Amazon,
let's take the question to the next level: Apply for the <span style="color:orange">orange</span> belt
in R: How you *wouldn't* solve the hanging cable problem by instead
solving the hanging cable problem **suspension bridge style**!

As explained in the video the
[catenary curve](https://en.wikipedia.org/wiki/Catenary) is the
[geometric shape](https://www.youtube.com/watch?v=npt6IkyL_f4&pbjreload=10),
a cable assumes under its own weight when supported only at its ends.
If instead the cable supports a uniformly distributed vertical load,
the cable has the shape of a
[parabolic curve](https://en.wikipedia.org/wiki/Parabola). This would
for example be the case for a
**[suspension bridge](https://en.wikipedia.org/wiki/Suspension_bridge)**
with a horizontal suspended deck, if the cable itself is not too
heavy compared to the road sections. A prominent example of a
suspension bridges is the
[Golden Gate Bridge](https://en.wikipedia.org/wiki/Golden_Gate_Bridge),
which we will use as motivating example for this post.

## Solving the Cable Problem

### Parabola Shape

Rephrasing the cable problem as the '*suspension bridge problem*' we
need to solve a two-component non-linear equation system:

1. the first component ensures that the parabolic curve with vertex at
$(0,0)$ goes through the poles at the x-values $-x$ and $x$. In other
words: the distance between the two poles is $2x$.  Note that the
coordinate system is aligned such that the lowest point of the cable
is at the origo.

2. the second component ensures that the arc-length of the parabola is
as given by the problem. Since the parabola is symmetric it is
sufficient to study the positive x-axis

The two criteria are converted into an equation system as follows:
$$
\begin{align*} a x^2 &= 50 - \text{height above ground} \\
\int_0^x \sqrt{1 + \left(\frac{d}{du} a u^2\right)^2} du &= 40.
\end{align*}
$$

Here, the general equation for
[arc-length](https://en.wikipedia.org/wiki/Arc_length) of a function
$y=f(u)$ has been used.  Solving the arc-length integral for a
parabola can either be done by numerical integration or by
[solving the integral analytically](http://www.math.drexel.edu/~tolya/arc_length_x%5e2.pdf)
or just look up the resulting analytic expression as eqn 4.25 in
@spiegel1968. Subtracting the RHS from each LHS gives us a non-linear equation
system with unknowns $a$ and $x$ of the form

$$
\left[
\begin{array}{c}
y_1(a,x) \\
y_2(a,x)
\end{array}
\right]
=
\left[
\begin{array}{c}
0 \\
0
\end{array}
\right].
$$

Writing this in R code:

```{r PARABOLA, echo=TRUE}
## Height of function at the location x from center is (pole_height - height_above_ground)
y1_parabola <- function(a, x, pole_height=50, above_ground=20) {
  a*x^2 - (pole_height - above_ground)
}

## Arc-length of the parabola between [-x,x] is given as cable_length
y2_parabola <- function(a,x, cable_length=80, arc_method=c("analytic","numeric")) {

  ##Arc-length of a parabola a*u^2 within interval [0,x]
  if(arc_method == "numeric") {
    f <- function(u) return( sqrt(1 + (2*a*u)^2))
    half_arclength <- integrate(f, lower=0, upper=x)$value
  } else if (arc_method=="analytic") {
    half_arclength <-  1/(4*a)*(2*a*x*sqrt(4*a^2*x^2+1) + asinh(2*a*x))
  }

  ##The equation: s = cable_length/2
  half_arclength - cable_length/2
}

## The non-linear equation system \bm{y}(\theta) = \bm{0}, where the LHS
## is given by a list with two components containing y_1(\theta) and y_2(\theta)
f_sys <- function(theta, y, pole_height=50, above_ground=20, cable_length=80, ...) {
  ##Parameters
  a <- theta[1]
  x <- exp(theta[2]) ##ensure x is positive

  c(y[[1]](a,x, pole_height=pole_height, above_ground=above_ground),
    y[[2]](a,x, cable_length=cable_length, ...))
}

##Helper function to transform theta parameter vector to (a,x)'
theta2ax <- function(theta) {
  c(a=theta[1], x=exp(theta[2]))
}
```

To ensure $x>0$ we re-parametrized the equations with $\theta_2 =
\log(x)$ and provide the function `theta2ax` to backtransform the result. We can now use the
[`nleqslv`](https://cran.r-project.org/web/packages/nleqslv/index.html)
package to solve the non-linear equation system using a one-liner:
```{r, echo=TRUE}
y_parabola <- list(y1_parabola, y2_parabola)
sol_parabola <- nleqslv(x=c(0.1,0.1),f_sys, y=y_parabola,  arc_method="analytic")
theta2ax(sol_parabola$x)
```
In other words, for a cable of length 80m the pole of a suspension bridge will be located `r sprintf("%.1fm",theta2ax(sol_parabola$x)[2])` from the origo, which means the two poles of the bridge will be `r sprintf("%.1fm",2*theta2ax(sol_parabola$x)[2])` apart, which is also the span of the bridge.

Using `arc_method="numeric"` instead of the analytic solution gives
```{r NUMERICALPARABOLA}
sol_parabola2 <- nleqslv(x=c(0.1,0.1),f_sys, y=y_parabola, arc_method="numeric")
theta2ax(sol_parabola2$x)
```

It is re-assuring to see that the numerical integration method yields
the same result as the analytic method. The analytic method has
mathematical beauty, the numerical method allows the data scientist to
solve the problem without diving into formula compendiums or geometry.

### Catenary Shape

Using the same code, but with the y-functions formulated for the
catenary case we obtain

```{r, echo=TRUE}
## Value of y=f(u) evaluated at u=x
y1_catenary <- function(a,x, pole_height=50, above_ground=20) {
  a * cosh(x/a) - a - (pole_height- above_ground)
}
## Arc-length condition
y2_catenary <- function(a,x, cable_length=80) {
  a * sinh(x/a) - cable_length/2
}

## Solve equation system
y_catenary <- list(y1_catenary, y2_catenary)
sol_catenary <- nleqslv(x=c(0.1,0.1),f_sys, y=y_catenary, method="Newton")
theta2ax(sol_catenary$x)
```

In other words the solution to the original cable problem is
$x=`r sprintf("%.1f m",theta2ax(sol_catenary$x)[2])`$ whereas the answer to
the suspension bridge version is $x=`r sprintf("%.1fm",theta2ax(sol_parabola$x)[2])`$.
The difference to the parabolic form can be seen from the following
graph:

```{r PARABOLACATENARYPLOT}
data.frame(x=seq(-25,25,length=1000)) %>%
  mutate(catenary=y1_catenary(x, a=sol_catenary$x[1], above_ground=50),
         parabola=y1_parabola(x, a=sol_parabola$x[1], above_ground=50)) %>%
  tidyr::gather("Shape","y", -x) %>%
  ggplot(aes(x=x,y=y, color=Shape)) + geom_line() + xlab("Distance from center [m]") + ylab("Height")
```

## Testing the theory

```{r}
##true height is 230m, see https://en.wikipedia.org/wiki/Golden_Gate_Bridge, but picture is distorted
gg_pole_height <- gg_pole_height <- 192
##true height above sealvl is 67.1m, but that's the bridge not the cable
gg_above_ground <- 75
##half the span of the bridge, total is 1280m,
gg_pole_position <- 640
```

We test our theory by studying the cable of the Golden Gate suspension
bridge.  Shown below is a photograph by
[D Ramey Logan](https://commons.wikimedia.org/wiki/File:Golden_Gate_Bridge_Dec_15_2015_by_D_Ramey_Logan.jpg)
available under a
[CC BY 4.0](https://creativecommons.org/licenses/by/4.0/deed.en)
license. For presentation in this post the image was tilted by -0.75
degrees (around the camera's view axis) with the
[`imager`](https://dahtah.github.io/imager/imager.html) package to
make sea level approximately horizontal. Parabolic and catenary
overlays (no real difference between the two) were done using the
theory described above.

```{r PREPROCESS, echo=TRUE}
##Preprocess image
img <- imager::load.image(file.path(fullFigPath, "Golden_Gate_Bridge.png"))
img <- imager::imrotate(img, angle=-0.75, interpolation=1)
img <- imager::resize(img,-50,-50, interpolation_type=5)
```
```{r, eval=FALSE}
##Fix negative channel values, if one uses interpolation=2 in imrotate
img[img < 0] <- 0
img[img > 1] <- 1
```
We manually identify center, sea level and poles from the image and
use [`annotation_raster`](https://ggplot2.tidyverse.org/reference/annotation_raster.html)
to overlay the image on the `ggplot` of the
corresponding parabola and catenary. See the [code](`r paste0("https://raw.githubusercontent.com/hoehleatsu/hoehleatsu.github.io/master/_source/",current_input())`)
on github for details.

```{r GOLDENGATE, fig.width=8, fig.height=2}
##Convert to ggplot compatible format (alternative: as.data.frame)
ggate <- aperm(img[,,1,], perm=c(2,1,3))

a_parabola <- uniroot(function(a) y1_parabola(a, x=gg_pole_position, pole_height=gg_pole_height, above_ground=gg_above_ground), interval=c(0, 30))$root
a_catenary <- uniroot(function(a) y1_catenary(a, x=gg_pole_position, pole_height=gg_pole_height, above_ground=gg_above_ground), interval=c(1e-12, 5000))$root

p <- data.frame(x=seq(-800,800,length=1000)) %>%
  mutate(catenary=y1_catenary(x, a=a_catenary, pole_height=gg_pole_height, above_ground=gg_above_ground) + gg_pole_height,
         parabola=y1_parabola(x, a=a_parabola, pole_height=gg_pole_height, above_ground=gg_above_ground) + gg_pole_height) %>%
  tidyr::gather("Shape","y", -x) %>%
  ggplot(aes(x=x,y=y, color=Shape)) +
#  geom_line() +
  ylim(c(-60, NA)) +
  scale_x_continuous(name="Distance from center [distorted m]", breaks=seq(-800,800,by=100)) +
  ylab("Height [distorted m]")

##This is a rather messy piece of code scaling the image to fit acceptably
##to known parameters. But since the photo is not really orthogonal to the bridge
##this only works very very approximately...
xy_ratio <- dim(ggate)[1] / dim(ggate)[2]
x_scale <- 1.7 ##ymin=-110
p + annotation_raster(ggate, ymin = -124,ymax= 2*gg_pole_position*xy_ratio,xmin = -x_scale*gg_pole_position, xmax = x_scale*gg_pole_position, interpolate=FALSE) +
  coord_equal() +
  geom_line(lwd=1, alpha=0.6) +
  geom_text(data=data.frame(x=0, y=200,text=paste0(a_parabola, " %.% x^2"),Shape=NA), aes(x=x, y=y, label=text),parse=TRUE, show.legend=FALSE) +
  geom_hline(yintercept=0,color="steelblue",lty=2, size=0.8) +
  geom_line(data=data.frame(x=rep(-gg_pole_position, 2), y=c(0,gg_pole_height), Shape=NA),show.legend=FALSE,lwd=1.2) +
  geom_line(data=data.frame(x=rep(gg_pole_position, 2), y=c(0,gg_pole_height), Shape=NA),show.legend=FALSE,lwd=1.2) +
  geom_point(data=data.frame(x=0, y=gg_above_ground, Shape=NA), show.legend=FALSE, size=1.5) +
  geom_line(data=data.frame(x=rep(0,2), y=c(0,gg_above_ground), Shape=NA), show.legend=FALSE, lty=2)
```

The fit is not perfect, which is due to the camera's direction not
being orthogonal to the plane spanned by the bridge -
for example the right pole appears to be closer to the camera than the
left pole^[A manual investigation using the "Map | Map Object" Filter in Gimp showed that the angle of tilting around the y-axis is about 20 degrees.]. We
scaled and 'offsetted' the image so the left pole is at distance `r gg_pole_position`m from origo, but did not correct for the tilting around the $y$-axis. Furthermore, distances are being distorted
by the lens, which might explain the poles being too
small. [Rectification](https://en.wikipedia.org/wiki/Image_rectification)
and
[perspective control](https://en.wikipedia.org/wiki/Perspective_control)
of such images is a
**[photogrammetric](https://en.wikipedia.org/wiki/Photogrammetry)**
method beyond the scope of this post!

## Discussion

This post may not to impress a Matlab coding engineer, but it shows
how R has developed into a versatile tool going way beyond statistics:
We used its optimization and image analysis capabilities. Furthermore,
given an analytic form of $y(\theta)$, R can symbolically determine
the Jacobian and, hence, implement the required Newton-Raphson solving
of the non-linear equation system directly - see the Appendix. In
other words: R is also a full stack mathematical problem solving tool!

As a **challenge** to the interested reader: Can you write R code,
for example using `imager`, which automatically identifies poles and cable in
the image and based on the known specification of these parameters of
the Golden Gate Bridge (pole height: 230m, span 1280m, clearance above
sea level: 67.1m), and perform a rectification of the image? If yes,
Stockholm University's Math Department
[hires](https://www.math.su.se/english/education/phd-studies/admission-and-vacant-positions)
for Ph.D. positions every April! The challenge could work well as pre-interview project. `r emo::ji("wink")`

## Appendix - Newton-Raphson Algorithm

Because the `y_1(a,x)` and `y_2(a,x)` are both available in closed
analytic form, one can form the Jacobian of non-linear equations
system by combining the two gradients. This can be achieved
symbolically using the `deriv` or
[`Deriv::Deriv`](https://cran.r-project.org/web/packages/Deriv/index.html)
functions in R.

Given starting value $\theta$ the iterative procedure to find the root
of the non-linear equation system $y(\theta) = 0$ is given by
[@nocedal_wright2006, Sect. 11.1]

$$
\theta^{(k+1)} = \theta^k - J(\theta^k)^{-1} y(\theta),
$$

where $J$ is the
[Jacobian](https://en.wikipedia.org/wiki/Jacobian_matrix_and_determinant)
of the system, which in this case is a 2x2 matrix.

```{r NEWTONRAPHSON, echo=TRUE}
gradient_y1 <- Deriv::Deriv(y1_parabola, x=c("a","x"))

y2_parabola_analytic <- function(a,x, cable_length=80) {
  1/(4*a)*(2*a*x*sqrt(4*a^2*x^2+1) + asinh(2*a*x)) - cable_length/2
}

gradient_y2 <- Deriv::Deriv(y2_parabola_analytic, x=c("a","x"))

##Jacobian
J <- function(theta, pole_height=50, above_ground=20, cable_length=80, ...) {
  a <- theta[1]
  x <- exp(theta[2]) #  x <- exp(theta[2])

  ##Since we use x = exp(theta[2])=g(theta[2]) we need the chain rule to find the gradient in theta
  ##this is g'(theta[2]) = exp(theta[2]) = x
  rbind(gradient_y1(a,x, pole_height=pole_height, above_ground=above_ground)* c(1, x),
        gradient_y2(a,x, cable_length=cable_length)  * c(1, x))
}
```

```{r JMANUAL, echo=FALSE, results='hide'}
J_manual <- function(theta) {

  a <- theta[1]
  x <- exp(theta[2])

  ##Derivatives dy_1/da and dy_1/dx found with maxima
  y1da <- x^2
  y2da <- (2*x*sqrt(4*a^2*x^2+1)+(8*a^2*x^3)/sqrt(4*a^2*x^2+1)+(2*x)/sqrt(4*a^2*x^2+1))/(4*a)-(asinh(2*a*x)+2*a*x*sqrt(4*a^2*x^2+1))/(4*a^2)


  ##Derivates dy_2/da and dy_2/dx found with maxima + chain rule
  y1dx <- 2*a*x * x
  y2dx <- (2*a*sqrt(4*a^2*x^2+1)+(8*a^3*x^2)/sqrt(4*a^2*x^2+1)+(2*a)/sqrt(4*a^2*x^2+1))/(4*a) * x


  ##Return result
  matrix(c(y1da,y1dx, y2da, y2dx), 2,2, byrow=TRUE)
}

##Start value
theta0 <- c(0.1,log(10))
##Compare Jacobian at theta0
J(theta=theta0)
J_manual(theta=theta0)
```

By iterating Newton-Raphson steps we can find the solution of the
equation system manually:

```{r NRITERATE, echo=TRUE}
##Start values
theta <- c(0.1,log(10))
thetanew <- c(0.1,log(20))
##Log with the values
log <- t(theta2ax(theta))

##Iterate Newton-Raphson steps until convergence
while ( (sum(thetanew - theta)^2 / sum(theta^2)) > 1e-15) {
  theta <- thetanew
  ##Update step
  thetanew <- theta - solve(J(theta=theta)) %*% f_sys(theta, y=y_parabola, arc_method="analytic")
  ##Add to log
  log <- rbind(log, theta2ax(thetanew))
}

##Look at the steps taken
log
```

We show the moves of the algorithm in a 2D contour plot for $r(a,x) =
\sqrt{y_1(a,x)^2 + y_2(a,x)^2}$. The solution to the system has
$r(a,x)=0$. See the [code](`r
paste0("https://raw.githubusercontent.com/hoehleatsu/hoehleatsu.github.io/master/_source/",current_input())`)
on github for details.

```{r NRSTEPSPLOT, warning=FALSE}
grid <- expand.grid(a=seq(0.01,0.1, length=50), x=seq(10,30, length=50)) %>% as.data.frame
##apply(grid, 2, fun) would be an alternative
grid %<>% rowwise %>%
  mutate(r = sqrt(sum( f_sys( c(a,log(x)), y=y_parabola, arc_method="analytic")^2))) %>%
  ungroup



ggplot(grid, aes(x=a, y=x, z=r)) +  geom_raster(aes(fill = r)) +
  viridis::scale_fill_viridis(option="magma",direction=-1) +
  geom_contour(colour = "white") +
  geom_segment(data=as.data.frame(log) %>% mutate(r=NA), aes(x=a, xend=lag(a), y=x, yend=lag(x)),arrow = arrow(length=unit(0.30,"cm"), ends="first", type = "closed"),color="chartreuse3") +
  geom_point(data=as.data.frame(log) %>% mutate(r=NA), color="salmon3") +
  geom_point(data=data.frame(a=theta2ax(sol_parabola$x)[1], x=theta2ax(sol_parabola$x)[2],r=NA), aes(x=a, y=x), pch=4, color="salmon3",size=2) +
  xlim(c(0.01,0.1)) +
  ylim(c(10,30))
```

## Literature

