---
layout: post
title: "World Income, Inequality and Murder"
tags: [rstats, stats, economics, data visualization, income, equality]
#  bibliography: ~/Literature/Bibtex/jabref.bib
header-includes:
   - \usepackage{bm}
comments: true
editor_options:
  chunk_output_type: console
---

```{r,include=FALSE,echo=FALSE,message=FALSE}
##If default fig.path, then set it.
if (knitr::opts_chunk$get("fig.path") == "figure/") {
  knitr::opts_knit$set( base.dir = '/Users/hoehle/Sandbox/Blog/')
  knitr::opts_chunk$set(fig.path="figure/source/2018-07-09-gini/")
}
fullFigPath <- paste0(knitr::opts_knit$get("base.dir"),knitr::opts_chunk$get("fig.path"))
filePath <- file.path("","Users","hoehle","Sandbox", "Blog", "figure", "source", "2018-07-09-gini")

knitr::opts_chunk$set(echo = TRUE,fig.width=8,fig.height=5,fig.cap='',fig.align='center',echo=FALSE,dpi=72*2) # autodep=TRUE
options(width=150)

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(knitr))

## Packages used for this post
suppressPackageStartupMessages(require(reldist))
suppressPackageStartupMessages(require(gridExtra))
suppressPackageStartupMessages(require(grid))
suppressPackageStartupMessages(require(scales))

##Configuration
options(knitr.table.format = "html")
theme_set(theme_minimal())
#if there are more than n rows in the tibble, print only the first m rows.
options(tibble.print_max = 10, tibble.print_min = 5)
```

## Abstract:

We follow up on last weeks post on using Gapminder data to study the
world's income distribution. In order to assess the inequality of the
distribution we compute the Gini coefficient for the world's income
distribution by Monte Carlo approximation and visualize the result as
a time series. Furthermore, we animate the association between Gini
coefficient and homicide rate per country using the new version of
`gganimate`.

<center>
```{r,results='asis',echo=FALSE,fig.cap="Gini coeAnimated Income Mountains"}
cat(paste0("![]({{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"MURDERGINIPLOT-1.png"),")")
```
</center>

{% include license.html %}


## Introduction

One of the main messages of the Chapter 'The Gap Instinct' of the book
[*Factfulness*](https://www.gapminder.org/factfulness/) is that there
is no justification of the 'we' and 'them' classification of countries
anymore, because 'they' have converged towards the same levels in key
indicators such as life expectancy, child mortality, births per
female. The difference between countries is, hence, not as many
imagine it to be: there is less inequality and no real gap. While
reading, I became curious about the following: **what if countries
became more equal, but simultaneously inequality within countries
became bigger?** This was also indirectly a Disqus comment by
F. Weidemann to the post
[*Factfulness: Building Gapminder Income Mountains*](http://staff.math.su.se/hoehle/blog/2018/07/02/factfulness.html).
Aim of the present post is to investigate this hypothesis using the
Gapminder data by calculating Gini coefficients. Furthermore, we use the country specific Gini coefficients to investigate the association with the number of homicides in the country.

```{r LOADDATA}
dataPath <- file.path(knitr::opts_knit$get("base.dir"),"figure/source/2018-07-02-factfulness")
load(file=file.path(dataPath, "factfulness.RData"))
```

## Gini coefficient

There are  different ways to measure income inequality, both in
terms of which response you consider and which statistical summary you
compute for it. Not going into the details of these discussion we
use the GDP/capita in
[Purchasing Power Parities (PPP)](https://en.wikipedia.org/wiki/Purchasing_power_parity)
measured in so called international dollars (fixed prices 2011). In
other words, comparison between years and countries are possible,
because the response is adjusted for inflation and differences in
price of living.

The
[**Gini coefficient**](https://en.wikipedia.org/wiki/Gini_coefficient)
is a statistical measure to quantify inequality. In what follows we
shall focus on computing the Gini coefficient for a continuous
probability distribution with a known probability density
function. Let the probability density function of the non-negative
continuous income distribution be defined by $f$, then the Gini
coefficient is given as **half the relative mean difference**:

$$
G
= \frac{1}{2\mu}\int_0^\infty \int_0^\infty |x-y| \> f(x) \> f(y) \>
dx\> dy, \quad\text{where}\quad \mu = \int_{0}^\infty x\cdot f(x) dx.
$$

Depending on $f$ it might be possible to
[solve these integrals analytically](https://en.wikipedia.org/wiki/Gini_coefficient#Continuous_probability_distribution),
however, a straightforward computational approach is to use Monte Carlo
sampling - as we shall see shortly. Personally, I find the above relative mean difference presentation of the Gini index much more intuitive than the area
argument using the Lorenz curve. From the eqution it also becomes
clear that the Gini coefficient is invariant to multiplicative
changes in the income: if everybody increases their income by factor
$k>0$ then the Gini coefficient remains the same, because $|k x - k y|
= k | x - y|$ and $E(k \cdot X) = k \mu$ and, hence, $k$ cancels from
numerator and denominator.


The above formula indirectly also states
how to compute the Gini coefficient for a discrete sample of size $n$
and with incomes
$x_1,\ldots, x_n$:
$$
G =  \frac{\sum_{i=1}^n \sum_{j=1}^n  |x_i -
x_j| \frac{1}{n} \frac{1}{n}}{2 \sum_{i=1}^n  x_i \frac{1}{n}} =
\frac{\sum_{i=1}^n \sum_{j=1}^n |x_i - x_j|}{2 n \sum_{j=1}^n x_j}.
$$


#### Approximating the integral by Monte Carlo

If one is able to easily sample from $f$ then can instead of solving
the integral analytically use $k$ pairs $(x,y)$ both drawn at random
from $f$ to approximate the double integral:

$$
G \approx \frac{1}{2\mu K} \sum_{k=1}^K |x_k - y_k|, \quad\text{where}\quad
x_k \stackrel{\text{iid}}{\sim} f \text{ and } y_k \stackrel{\text{iid}}{\sim} f,
$$
where for our mixture model
$$
\mu = \sum_{i=1}^{`r n_country`} w_i \> E(X_i) = \sum_{i=1}^{`r n_country`} w_i \exp\left(\mu_i + \frac{1}{2}\sigma_i^2\right).
$$
This allows us to compute $G$ even in case of a complex $f$ such as the log-normal mixture distribution. As always, the larger $K$ is, the better the Monte Carlo approximation is.

```{r MONTECARLO, cache=TRUE, echo=TRUE}
##Precision of Monte Carlo approx is controlled by the number of samples
K <- 1e6

##Compute Gini index of world income per year
gini_year <- gm %>% group_by(year) %>% do({
  x <- rmix(K, meanlog=.$meanlog, sdlog= .$sdlog, w=.$w)
  y <- rmix(K, meanlog=.$meanlog, sdlog= .$sdlog, w=.$w)
  int <- mean( abs(x-y) )
  mu <- sum(exp( .$meanlog + 1/2 * .$sdlog^2) * .$w)
  data.frame(gini_all_mc=1/(2*mu)*int,
             country_gini=gini(.$gdp*.$population))
}) %>%
  rename(`World Population GDP/capita`=gini_all_mc, `Country GDP/capita`=country_gini)

##Convert to long format
gini_ts <- gini_year %>% gather(key="type", value="gini", -year)
```

### Results

We start by showing the country specific Gini coefficient per year since 1950 for a
somewhat arbitrary selection of countries.
The dashed black line shows the mean Gini coefficient each year over
all `r n_country` country specific Gini coefficients in the dataset.

```{r GINITSSELECTED, fig.height=4}
##Mean of all ginis per year
Mean <- gm %>% group_by(year) %>% summarise(country="Mean", gini=mean(gini))

countryList <- c("Japan", "United States","Sweden", "Germany", "China","India", "Russia","Brazil", "Nigeria", "Ethopia")
gm %>% filter(country %in% countryList, year >= 1950) %>%
  rename(Country = country) %>%
  ggplot(aes(x=as.numeric(year), y=gini, color=Country)) + geom_line(lwd=1.1) +
  geom_line(data=Mean %>% filter(year >= 1950), color="black", lty=2, lwd=1.1) +
  xlab("Time [Years]") + ylab("Gini coefficient") +
  scale_y_continuous(labels=scales::percent) +
  scale_x_continuous(breaks=seq(1950,2015,by=10))
```

In addition, we compute and show as time series
the Gini coefficient of the `r n_country`
countries' GDP/capita per year. Furthermore, we show the Monte Carlo computed Gini coefficient for
the world's income distribution given as a log-normal mixture with
`r n_country` components.

```{r PLOTGINI}
##Plot the mean gini coeff over time
plot_gini_ts <- function(from=1800, ylim=c(0,1), n_y_ticks=5) {
  ##Restrict to time range
  ts <- gini_ts %>% filter(year>=from)

  ##Make the plot
  p <- ggplot(ts, aes(x=as.numeric(year), y=gini, color=type)) +
    geom_line(lwd=1.2) +
    xlab("Time [year]") + ylab("Gini coefficient [in %]") +
    scale_y_continuous(labels=scales::percent,breaks = seq(0,1, by=1/n_y_ticks),limit=ylim) +
    ggtitle(paste0("Gini coefficient for GDP/capita (PPP$, 2011)")) +
    theme(legend.position="bottom") +
    guides(color=guide_legend(title="")) +
    theme(plot.margin = unit(c(1,1,1,1),"cm")) +
    annotation_custom(grob = textGrob("Data Source: gapm.io/d_gini",gp = gpar(fontsize = 7, col = "gray40"),x=1, just=c("right","centre"), y=-0.18)) +
    annotation_custom(grob = textGrob("Created by: m_hoehle",gp = gpar(fontsize = 7, col = "gray40"),x=1, just=c("right","centre"), y=-0.22))
  gt <- ggplotGrob(p)
  gt$layout$clip[gt$layout$name=="panel"] <- "off"

  gt

  grid.draw(gt)
  invisible(gt)
}

plot_gini_ts(from=1800)
```

We notice that the Gini coefficient for the `r n_country`
countries' GDP/capita remains very stable over time. This, however,
does not take the large differences in populations between countries
into account. A fairer measure is thus the Gini coefficient for the
world's income distribution. We see that this Gini coefficient
increased over time until peaking around 1990. From then on it has
declined.  However, the pre-1950 Gini coefficients are rather
guesstimates as stated by Gapminder, hence, we zoom in on the period
from 1970, because data are more reliable from this point on.

```{r PLOTGINIFROM1970}
plot_gini_ts(from=1970, ylim=c(NA,NA), n_y_ticks=20)
```

## Gini coefficient and Homicide Rate

```{r MURDERDATA}
if (!file.exists(file.path(fullFigPath,"murders.xlsx"))) {
  download.file("https://docs.google.com/spreadsheet/pub?key=tZgPgT_sx3VdAuyDxEzenYA&output=xlsx", file.path(fullFigPath,"murders.xlsx"))
}

##Fetch murder rates per country and convert to long-format
murder <- readxl::read_xlsx(file.path(fullFigPath,"murders.xlsx")) %>%
  rename("country"="Murder per 100,000, age adjusted") %>%
  gather(key="year", value="murder_rate", -country) %>%
  mutate(year=as.numeric(year)) %>% filter(!is.na(murder_rate))

##Join to gm dataset
gm2 <- gm %>% mutate(year=as.numeric(year)) %>%
  inner_join(murder, by=c("country", "year")) %>% rename(Region=region)

##Massage so we fix zeroes (not possible to do log10(0))
gm2_nozero <- gm2 %>% mutate(murder_rate=if_else(murder_rate == 0, 0.01, murder_rate))

##Show year
show_year <- 2005
```

Finally, we end the post by illustrating the association between the
Gini coefficient and the homicide rate per country using a 2D
scatterplot over the years. The Gapminder data download page also
contains
[data](https://docs.google.com/spreadsheet/pub?key=tZgPgT_sx3VdAuyDxEzenYA&output=xlsx)
for this for the years `r min(murder$year)`- `r max(murder$year)`.
Unfortunately, no data for more recent years are available from the
Gapminder homepage, but the plot shown below is the situation in
`r show_year` with a log-base-10 y-axis for the homicide rates.  For each
of the four Gapminder regional groups we also fit a simple linear
regression line to the points of all countries within the region.
Findings such as @fajnzylber_etal2002 suggest that there is a strong
positive correlation between Gini coefficient and homicide
rate. To illustrate this the thin dashed line is the result of a
linear regression (on the log-base-10 scale) for all data
points. However, we see from the plot that there are regional
differences even having a reversed sign of the relationship. Of course
correlation does not imply causality and explanations for this
relationship are
[debated](https://www.theguardian.com/us-news/2017/dec/08/income-inequality-murder-homicide-rates)
and beyond the scope of this post.

```{r MURDERGINIPLOT}
p <- ggplot(gm2_nozero %>% filter(year==show_year), aes(x=gini, y=murder_rate,size=population, color=Region)) +
  geom_point() +
  scale_x_continuous(labels=scales::percent) +
  scale_y_continuous(trans="log10",
                     breaks = trans_breaks("log10", function(x) 10^x,n=6),
                     labels = trans_format("log10", function(x) ifelse(x<0, sprintf(paste0("%.",ifelse(is.na(x),"0",round(abs(x))),"f"),10^x), sprintf("%.0f",10^x)))) +
  geom_smooth(aes(x=gini, y=murder_rate), se=FALSE, method="lm", formula=y~x, color="lightgray",lty=2,lwd=0.5) +
  geom_smooth(se=FALSE, method="lm", formula=y~x) +
  geom_text(data=gm2 %>% filter(year == show_year), aes(x=gini,y=murder_rate, label=country), vjust=-0.9, show.legend=FALSE) +
  ylab("Murder rate [per 100,000 population]") +
  xlab("Gini coefficient [in %]") +
  guides(size=FALSE)
print(p)
```

We extend the plots to all years `r min(murder$year)`-`r max(murder$year)`.  Unfortunately,
not all countries are available every year - so we only
plot the available countries each year. This means that many African countries are
missing from the animation. An improvement would be to try some form
of linear interpolation. Furthermore, for the sake of simplicity of illustration,
we fix countries with a reported murder rate of zero in a given year
(happens for example for Cyprus, Iceland, Fiji in some years) to 0.01
per 100,000 population. This can be nicely animated using the new version
of the [`gganimate`](https://github.com/thomasp85/gganimate) package by
[Thomas Lin Pedersen](https://twitter.com/thomasp85).


```{r ANIMATE, message=FALSE, warning=FALSE, echo=TRUE}
## New version of gganimate. Not on CRAN yet.
## devtools::install_github('thomasp85/gganimate')
require(gganimate)

p <- ggplot(gm2_nozero, aes(x=gini, y=murder_rate,size=population, color=Region)) +
  geom_point() +
  scale_x_continuous(labels=scales::percent) +
  scale_y_continuous(trans="log10",
                     breaks = trans_breaks("log10", function(x) 10^x,n=5),
                     labels = trans_format("log10", function(x) ifelse(x<0, sprintf(paste0("%.",ifelse(is.na(x),"0",round(abs(x))),"f"),10^x), sprintf("%.0f",10^x)))) +
  geom_smooth(se=FALSE, method="lm", formula=y~x) +
  geom_text(data=gm2, aes(x=gini,y=murder_rate, label=country), vjust=-0.9, show.legend=FALSE) +
  ylab("Murder rate [per 100,000 population]") +
  xlab("Gini coefficient [in %]") +
  guides(size=FALSE) +
  labs(title = 'Year: {frame_time}') +
  transition_time(year) +
  shadow_wake(wake_length=0.15, exclude_layer=c(2,3)) +
  ease_aes('linear')

animate(p, nframes=length(unique(gm2$year)), fps=4, width=800, height=400, res=100)
```

## Discussion

Based on the available Gapminder data we showed that in the last 25
years the Gini coefficient for the world's income distribution has
decreased. For several individual countries opposite dynamics are,
however, observed. One particular concern is the share that the
richest 1% have of the overall wealth:
[more than 50%](https://www.theguardian.com/inequality/2017/nov/14/worlds-richest-wealth-credit-suisse).

```{r REDDITPLOTS, results='hide'}
##Gini vs. murder rate for reddit
p <- ggplot(gm2_nozero %>% filter(year==show_year), aes(x=gini, y=murder_rate,size=population, color=Region)) +
  geom_point() +
  scale_x_continuous(labels=scales::percent) +
  scale_y_continuous(trans="log2",
                     breaks = trans_breaks("log2", function(x) 2^x,n=6),
                     labels = trans_format("log2", function(x) ifelse(x<0, sprintf(paste0("%.",ifelse(is.na(x),"0",round(abs(x))),"f"),2^x), sprintf("%.0f",2^x)))) +
  geom_smooth(aes(x=gini, y=murder_rate), se=FALSE, method="lm", formula=y~x, color="lightgray",lty=2,lwd=0.5) +
  geom_smooth(se=FALSE, method="lm", formula=y~x) +
  geom_text(data=gm2 %>% filter(year == show_year), aes(x=gini,y=murder_rate, label=country), vjust=-0.9, show.legend=FALSE) +
  ylab("Murder rate [per 100,000 population]") +
  xlab("Gini coefficient [in %]") +
  guides(size=FALSE) +
  theme(legend.position="bottom") +
  ggtitle(paste0("Scatterplot of year 2005 Gini coefficient for Income vs. Murder Rate")) + 
  annotation_custom(grob = textGrob("Data Source: gapm.io/d_gini",gp = gpar(fontsize = 7, col = "gray40"),x=1, just=c("right","centre"), y=0.05)) +
  annotation_custom(grob = textGrob("Created by: m_hoehle",gp = gpar(fontsize = 7, col = "gray40"),x=1, just=c("right","centre"), y=0.03)) 
gt <- ggplotGrob(p)
gt$layout$clip[gt$layout$name=="panel"] <- "off"

grid.draw(gt)
ggsave(file=file.path(fullFigPath,"murdergini.png"), width=8,height=6, dpi=300)


##GDP/capita vs. murder rate for reddit
p <- ggplot(gm2_nozero %>% filter(year==show_year), aes(x=gdp, y=murder_rate,size=population, color=Region)) +
  geom_point() +
  scale_x_continuous(trans="log2",
                     breaks = trans_breaks("log2", function(x) 2^x,n=6),
                     labels = trans_format("log2", function(x) ifelse(x<0, sprintf(paste0("%.",ifelse(is.na(x),"0",round(abs(x))),"f"),2^x), sprintf("%.0f",2^x)))) +
  scale_y_continuous(trans="log2",
                     breaks = trans_breaks("log2", function(x) 2^x,n=6),
                     labels = trans_format("log2", function(x) ifelse(x<0, sprintf(paste0("%.",ifelse(is.na(x),"0",round(abs(x))),"f"),2^x), sprintf("%.0f",2^x)))) +
  geom_smooth(aes(x=gdp, y=murder_rate), se=FALSE, method="lm", formula=y~x, color="lightgray",lty=2,lwd=0.5) +
  geom_smooth(se=FALSE, method="lm", formula=y~x) +
  geom_text(data=gm2 %>% filter(year == show_year), aes(x=gdp,y=murder_rate, label=country), vjust=-0.9, show.legend=FALSE) +
  ylab("Murder rate [per 100,000 population]") +
  xlab("GDP/capita [PPP$, fixed 2011 price]") +
  guides(size=FALSE) +
  theme(legend.position="bottom") +
  ggtitle(paste0("Scatterplot of year 2005 GDP/capita vs. Murder Rate")) + 
  annotation_custom(grob = textGrob("Data Source: gapm.io/d_gini",gp = gpar(fontsize = 7, col = "gray40"),x=0, just=c("left","centre"), y=0.05)) +
  annotation_custom(grob = textGrob("Created by: m_hoehle",gp = gpar(fontsize = 7, col = "gray40"),x=0, just=c("left","centre"), y=0.03)) 
gt <- ggplotGrob(p)
gt$layout$clip[gt$layout$name=="panel"] <- "off"

grid.draw(gt)

ggsave(file=file.path(fullFigPath,"murdergdp.png"), width=8,height=6, dpi=300)
```

## Literature

