---
layout: post
title: "Purr yourself into a math genius"
tags: [rstats, purrr, combinatorics, math puzzle]
#  bibliography: ~/Literature/Bibtex/jabref.bib
header-includes:
   - \usepackage{bm}
comments: true
editor_options:
  chunk_output_type: console
---

```{r,include=FALSE,echo=FALSE,message=FALSE}
##If default fig.path, then set it.
if (knitr::opts_chunk$get("fig.path") == "figure/") {
  knitr::opts_knit$set( base.dir = '/Users/hoehle/Sandbox/Blog/')
  knitr::opts_chunk$set(fig.path="figure/source/2019-01-04-mathgenius/")
}
fullFigPath <- paste0(knitr::opts_knit$get("base.dir"),knitr::opts_chunk$get("fig.path"))
filePath <- file.path("","Users","hoehle","Sandbox", "Blog", "figure", "source", "2019-01-04-mathgenius")

knitr::opts_chunk$set(echo = TRUE,fig.width=8,fig.height=4,fig.cap='',fig.align='center',echo=FALSE,dpi=72*2)#, global.par = TRUE)
options(width=150, scipen=1e3)

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(knitr))

##Configuration
options(knitr.table.format = "html")
theme_set(theme_minimal())
#if there are more than n rows in the tibble, print only the first m rows.
options(tibble.print_max = 10, tibble.print_min = 5)
```

## Abstract:

We use the purrr package to solve a popular math puzzle via a
combinatorial functional programming approach. A small shiny app is
provided to allow the user to solve their own variations of the
puzzle.

<center>
```{r,results='asis',echo=FALSE,fig.cap=""}
cat(paste0("<img src=\"{{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"shinyapp.png\" width=\"450\">\n"))
```
</center>


{% include license.html %}


## Introduction

[No. 4](http://www.briddles.com/2011/12/maths-puzzle.html) of the
[Top 5 hard math puzzles](http://www.briddles.com/riddles/top-5-hard-math)
at [briddles.com](www.briddles.com) goes like this:

<div class = "blackbox"> *How can I get the answer 24 by only using
the numbers 8,8,3,3.  You can use the main signs add, subtract
multiply and divide.*</div> <p>

Note: a solution has to use each of the specified 4 numbers exactly
ONCE, but they can be used in any order. In other words the standard
scheme is to solve expressions of the kind:

<center>
`a op1 b op2 c op3 d`
</center>
<p>
where `a`, `b`, `c` and `d` denote a permutation of the numbers
8, 8, 3, 3 and each of `op1`, `op2` and `op3` denotes the use of one binary
operator selected from +, -, * or /. An example is the expression
`8 + 3 + 8 * 3`. Parentheses are used to control the order in which the operators are applied, i.e.
`(8 + 3 + 8) * 3` yields a different result than `8 + 3 + (8 * 3)`.

After a few unsuccessful attempts to solve the above puzzle with pen
and paper it felt more *efficient* and computationally *challenging*
to solve this puzzle via a combinatorial approach: Simply
try out all permutations of the 4 numbers, the 3 binary operators and
all possible sets of parentheses to combine the operators. One can show
that there are at most

$$
\begin{align*} && \text{# permutations of the
$k=4$ base numbers} \\ \times && \text{# ways to select with replacement $(k-1)$ binary operators from the set $\{+,-,*,/\}$ }\\ \times &&
\text{# ways to parenthesize the $(k-1)$ binary operators} \\ &=&k!
\times 4^{(k-1)} \times \frac{1}{k} \binom{2k-2)}{k-1}
\end{align*}
$$
```{r, echo=FALSE}
catalan <- function(n) {
  choose(2*n, n) / (n+1)
}

nCombinations <- function(k) {
  factorial(k) * 4^(k-1) * catalan(k)
}
```

different combinations to choose from [^1]. As an example: for $k=4$ the maximal number of unique combinations is `r nCombinations(k=4)`.


#### Strategy

We will use a functional approach to solve the above combinatorial
problem. Why?

* because it seems like a good use-case for
[functional programming](https://en.wikipedia.org/wiki/Functional_programming),
* because it is important to extend your programming horizon every once in a
while, and
* because the
[`purrr`](https://cran.r-project.org/web/packages/purrr/index.html)
functional programming toolkit for R allows you to experiment with
this without having to leave the R universe [^2].

For those not
familar with `purrr` can find a wonderful didactic introduction in the
[useR! 2017 tutorial](https://github.com/cwickham/purrr-tutorial) by
[Charlotte Wickham](https://twitter.com/CVWickham).  Furthermore, learning
`purrr` was the 7th most frequent mentioned package in the
[#rstats users' 2019 R goals](https://masalmon.eu/2019/01/01/r-goals/).
In other words: Attention #rstats new years resolution makers: reading
this post is as **obligatory** as going to the gym on 01 Jan!

## Solving the Math Puzzle

We will divide-and-conquer the solution along the lines of the number
of combinations formula: Firstly, we will store all permutations of
the $(k-1)$ base numbers in a list `perm`. Secondly, we will store all
possible combinations of the $(k-1)$ operators in a list `operators`
and, thirdly, we generate all possible ways of putting parentheses around
the operators into a list `brackets`. Subsequently, we form the Cartesian
product of these three lists and build the corresponding expression for
each triple of permutation, operators and parentheses. Finally, each generated
expression is evaluated. The entire result is a data frame containing
all possible expressions and their associated value obtained when
evaluating the expression.

### Permutations of the base numbers

We let the variable `base_numbers` contain the specification of the
numbers to permute for the expression. The code should be written
general enough so it is possible to use a different base, e.g., $k=3$
or $k=5$.

```{r, echo=TRUE}
base_numbers <- c(8,8,3,3)
k <- length(base_numbers)

number_perm <- combinat::permn(base_numbers) %>%
  map(setNames, nm=letters[seq_len(k)])

##Slim in case permutations of the base numbers contain duplicates.
perm <- number_perm[!duplicated(map(number_perm, paste0, collapse=""))]
```

For $k=4$ this makes a total `r nCombinations(k=4)` combinations. However, since
the numbers 8 and 3 both appear more than once in the base numbers, we can slim
the number of permutations from `r length(combinat::permn(base_numbers))` to `r length(combinat::permn(base_numbers))/2/2`. Hence, there are altogether only `r nCombinations(k=4)/4` combinations to investigate.

### Combinations of the operators

The next step is to make all combinations of the $k-1$ binary operators
needed to combine the $k$ numbers. We use the string
format to represent the operators [^3] and thus just need the $k-1$'th
Cartesian product of the set $\{+, -, *, /\}$ represented as strings.
<!-- i.e. $\times_{i=1}^{k-1} \{+,-,*,/\}$. -->


```{r, echo=TRUE}
opList <- list("+", "-", "*", "/")
##Repeat the opList k-1 times
opsList <- map( seq_len(k-1), ~ opList)
##Form the Cartesian product
operators <- cross(opsList) %>%
  map( setNames, nm=paste0("op",seq_len(k-1)))
```

### Arrangements of the parentheses

As all the involved operators are binarym it becomes clear that
finding all possible ways to parenthesize the expression corresponds to finding all
[binary trees](https://en.wikipedia.org/wiki/Binary_tree) with $k-1$
leaves. Beautiful recursive code
inspiration for how to solve this can be found on
[leetcode.com](https://leetcode.com/problems/all-possible-full-binary-trees/solution/). Some
adaptation to R and our problem at hand was necessary - the idea is to
use recursion in $k$ and use a hash-map to cache results of previous
computations.

```{r, echo=TRUE}
##Initialize hashmap to save the results of all binary trees up to n=1 leaves
trees <- list()
trees[["0"]] <- NULL
trees[["1"]] <- list(list(val="node", left=NULL, right=NULL))
```

The rather elegant **recursive solution** for generating all binary trees
with $n$ leaves works by combining all possible ways to generate
subbranches containing $x$ and $n-x$ leaves, respectively:

```{r, echo=TRUE}
allBinTrees <- function(n) {
  ##Character version of n, which is used as hash key
  n_char <- as.character(n)

  ##Only compute something if n is not already in the hashlist.
  if (is.null(pluck(trees, n_char))) {
    trees[[n_char]] <<- list()

    ##Combine all possible ways  to generate bintrees with $i$ and $n-i$ leaves
    for (i in 1:(n-1)) {
      j = n - i
      for (left_tree in allBinTrees(i)) {
        for (right_tree in allBinTrees(j)) {
          trees[[n_char]][[length(trees[[n_char]]) + 1]] <<- list(val=NULL, left=left_tree, right=right_tree)
        }
      }
    }
  } #end if not already in tree list
  ##Return result from our hashmap
  return(pluck(trees, n_char))
}
```
We can test the function for $n=2$, which yields exactly one tree:

```{r TREE2STRING}
##Helper function to print a binary tree (recursive edition)
tree2String <- function(tree) {
  if (is.character(tree$val)) return(tree$val)
  paste0("(", tree2String(tree$left), " op " , tree2String(tree$right), ")")
}
```
```{r OPNUMHELP}
##Convention: Number the operators from left to right. We do the search
##and replace recursively. Any clever way to do this as a regexp?
addOpNumbers <- function(str, i=1) {
  if (!grepl(" op ", str)) return(str)
  ##Replace one "op"
  addOpNumbers( str=sub(" op ", paste0(" op",i," "), str), i=i+1)
}

##Convert the "node" placeholders into the variables a, b, c, ...
##Convention: Name the numbers from left to right by "a", "b", "c", ...
replaceNodes <- function(str, i=1) {
  if (!grepl("node", str)) return(str)
  ##Replace one "node"
  replaceNodes( str=sub("node", letters[i], str), i=i+1)
}
```

```{r TESTALLBINTREESN2, echo=TRUE}
##Manual construction
trees2 <- list(list(val=NULL, left=trees[["1"]][[1]], right=trees[["1"]][[1]]))
all.equal(allBinTrees(n=2), trees2)
```

The result is:
```{r TESTRESULT, echo=TRUE}
tree2String(allBinTrees(n=2)[[1]]) %>% replaceNodes() %>% addOpNumbers
```

In the above code segments the function `tree2String` is a small
helper function to convert the nested list structure to a string - in
this case: `` `r tree2String(allBinTrees(n=2)[[1]])` ``. Furthermore,
the function `replaceNodes` renames the terms `node` into the variables `` `r tree2String(allBinTrees(n=2)[[1]]) %>% replaceNodes()` ``. The `op`-strings are converted into numbered `op`-strings using `addOpNumbers`, i.e. the result becomes
`` `r tree2String(allBinTrees(n=2)[[1]]) %>% replaceNodes() %>% addOpNumbers` ``.
Details about the helper functions can be found in the [code](`r paste0("https://raw.githubusercontent.com/hoehleatsu/hoehleatsu.github.io/master/_source/",current_input())`)
on github.

With all preparations in place we can now generate all `r catalan(3)` possible ways to parenthesize the 3 binary operators using the following code:

```{r, echo=TRUE}
##Make all possible brackets
bracketing <- map_chr( allBinTrees(n=k),
                      ~ tree2String(.x) %>% addOpNumbers %>% replaceNodes)
```
```{r}
bracketing
```

### Putting it all together

We can now generate all combinations of numbers, operators and bracketing by the Cartesian of the three lists:

```{r, echo=TRUE}
combos <- cross3( perm, map( operators, unlist), bracketing) %>%
  map(setNames, c("numbers", "operators", "bracket"))
```

```{r}
##Even more general helper function for numbers as well as operators
replace <- function(str, what) {
  if (length(what) == 0) return(str)
  replace( str=gsub(names(what)[1], what[1],  str), what=what[-1])
}
```

We can now finally evaluate each of the `r length(combos)`
combinations. Note: Because this might take a while it's a good idea to add
a [progress bar](https://github.com/tidyverse/purrr/issues/149#issuecomment-359236625) for this `purrr` computation.

```{r, echo=TRUE}
##Set up a progress bar for use with the map function
pb <- progress_estimated(length(combos))

##Compute
res <- map(combos, .f=function(l) {
  pb$tick()$print()
  l[["expr"]] <- l[["bracket"]] %>% replace(l[["numbers"]]) %>% replace(l[["operators"]])
  l[["value"]] <- eval(parse(text=l[["expr"]]))
  return(l)
})
```
Again, `replace(v)` is a small helper function to replace the strings
in `names(v)` with `v`'s content. The actual evaluation of each
possible solution string is done by parsing the string with `parse`
and then evaluate the resulting expression. We extract the relevant results into a `data.frame`

```{r CONVERTTODF, warning=FALSE, echo=TRUE}
df <- map_df(res, ~ data.frame(expr=.x$expr, value=.x$value))
```

```{r DATATABLE}
DT::datatable(df)
```

We can now easily extract the solution:

```{r EXTRACT24, echo=TRUE}
##First element to give the value 24
detect(res, ~ isTRUE(all.equal(.x$value, 24)))
```

Voila! QED!

```{r}
solveMathPuzzle <- function(base_numbers, expr_result, operatorList) {

  ##Variables
  k <- length(base_numbers)

  ##Permuations of the base numbers
  perm <- combinat::permn(base_numbers) %>%
    map(setNames, nm=letters[seq_len(k)])

  ##Slim it?
  perm <- perm[!duplicated(map(perm, paste0, collapse=""))]

  ##Make all combinations of the operators
  opsList <- map( seq_len(k-1), function(.x) operatorList)
  operators <- cross(opsList) %>% map( setNames, nm=paste0("op",seq_len(k-1)))

  ##Make all possible brackets
  bracketing <- map_chr( allBinTrees(n=k),
                         ~ tree2String(.x) %>% addOpNumbers %>% replaceNodes)

  ##All combinations of the numbers, the order and the bracketing.
  ##Depending on the combinations this might take a while...
  combos <- cross3( perm, map( operators, unlist), bracketing) %>%
    map(setNames, c("numbers", "operators", "bracket"))

  ##Compute value of all combinations (with progress bar)
  ##Set up a progress bar for use with the map function
  pb <- progress_estimated(length(combos))

  ##Compute
  res <- map(combos, .f=function(l) {
    pb$tick()$print()
    l[["expr"]] <- l[["bracket"]] %>% replace(l[["numbers"]]) %>% replace(l[["operators"]])
    l[["value"]] <- eval(parse(text=l[["expr"]]))
    return(l)
  })

  ##Convert results to a data.frame
  pb <- progress_estimated(length(res))
  df <- map_df(res, ~ { pb$tick()$print() ; data.frame(expr=.x$expr, value=.x$value)})

  ##Match 24
  is_zero <- function(x) isTRUE(all.equal(x, 0))

  ##Only those with nice integers results
  df_int <- df %>%
    mutate(rounded = round(value, digits=0), diff=value - rounded) %>% rowwise %>%
    filter(is_zero(diff))

  res <- list(combos=df_int %>% select(expr, value),
              expr=df_int %>% filter(rounded==expr_result) %$% expr)

  return(res)
}
```

#### Extended New Years Fun
For user experimentation we wrapped all the above steps into one function `solveMathPuzzle` (see [github code](`r paste0("https://raw.githubusercontent.com/hoehleatsu/hoehleatsu.github.io/master/_source/",current_input())`) for details).
To underline the generalizability of the approach we solve a classical
2019 new-year's puzzle:

```{r COMPUTE2019, cache=TRUE, echo=TRUE}
res <- suppressWarnings(solveMathPuzzle( base_numbers=c(7,7,11,11,43,43), expr_result=2019, operatorList=c("+","*")))
res$expr[[1]]
```


## Shiny App

To make the above solution accessible to a wider audience we wrote a small
Shiny app to play with the code for $k=4$:

<center>
<!-- [https://hoehle.shinyapps.io/mathgenius/](https://hoehle.shinyapps.io/mathgenius/) -->
<a href="http://michaelhoehle.eu/shiny/mathgenius/">http://michaelhoehle.eu/shiny/mathgenius/</a>
</center>
<p>
<p>

Here one can alter the input numbers in case variants of the puzzle
are in need of a solution or, if you occasionally need to generate
math puzzles for your nephew...

<center>
```{r,results='asis',echo=FALSE,fig.cap=""}
cat(paste0("<img src=\"{{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"shinyapp.png\" width=\"600\">\n"))
```
</center>

Besides possible solutions one can view the result of all possible combinations yielding integer results in the "Details" tab. We invite you to experiment with the app or download the [source code of the Shiny app](`r paste0("https://raw.githubusercontent.com/hoehleatsu/hoehleatsu.github.io/master/figure/source/",gsub("\\.Rmd","",current_input()),"/mathgenius/app.R")`) from github for the full math experience. `r emo::ji("smiley")`

As always: it's amazing how easy you can wrap a interactive web based
UI around your running R code with Shiny!

## Discussion

We used a brute force solution approach by trying out all possible
combinations to solve the math puzzle. The code of our solution
approach is flexible enough to handle more or less base numbers,
however, the number of combinations to try quickly exceeds reasonable
memory and timing constraints. We stress that **a mathematical purr does
not need speed, it lives from the beauty of recursion and mappings**!
Clever mathematicians might be able to achieve considerable speed gains
by exploiting for example commutative properties of the operators whereas skilled computer scientists would parallelise the computations.


[^1]: Note: The term $\frac{1}{k} \binom{2k-2)}{k-1}$ is the so called
[Catalan number](https://en.wikipedia.org/wiki/Catalan_number#Applications_in_combinatorics),
which - among other applications - also denotes the number of ways to
parenthesize $k-1$ binary operators.

[^2]: Actually, the package more or less adds a lot of convenience wrapping for functional programming in R, the functional programming approach is rather deeply rooted in R due to the [S language being inspired by Scheme](https://web.archive.org/web/20140414081950/https://daspringate.github.io/posts/2013/05/Functional_programming_in_R.html).

[^3]: A purer functional approach would have been to use the function definition of the operators directly, i.e. to define `operatorList` with elements such as `` `+`(e1, e2) `` and then use these functions to build the parse tree as an expression. The disadvantage of such an approach is that the expressions become more cumbersome to write. For example `(5 + 3 + 2) * 4` as `` `*`( `+`( `+`(5, 3), 2), 4) ``.

## Literature
