---
layout: post
title: "Speedmining the Cubing Community with dbplyr"
tags: [rstats, dataviz, data science, tidyverse, Rubik's cube, 3x3x3, speedsolving]
#  bibliography: ~/Literature/Bibtex/jabref.bib
header-includes:
   - \usepackage{bm}
comments: true
editor_options:
  chunk_output_type: console
---

```{r,include=FALSE,echo=FALSE,message=FALSE}
##If default fig.path, then set it.
if (knitr::opts_chunk$get("fig.path") == "figure/") {
  knitr::opts_knit$set( base.dir = '/Users/hoehle/Sandbox/Blog/')
  knitr::opts_chunk$set(fig.path="figure/source/2019-05-06-wcamining/")
}
fullFigPath <- paste0(knitr::opts_knit$get("base.dir"),knitr::opts_chunk$get("fig.path"))
filePath <- file.path("","Users","hoehle","Sandbox", "Blog", "figure", "source", "2019-05-06-wcamining")

knitr::opts_chunk$set(echo = TRUE,fig.width=8,fig.height=4,fig.cap='',fig.align='center',echo=FALSE,dpi=72*2)#, global.par = TRUE)
options(width=150, scipen=1e3)

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(viridis))

##Configuration
options(knitr.table.format = "html")
theme_set(theme_minimal())
#if there are more than n rows in the tibble, print only the first m rows.
options(tibble.print_max = 10, tibble.print_min = 5)
```

## Abstract:

We use the `RMariaDB` and `dbplyr` packages to analyze the results
database of the World Cubing Association. In particular we are
interested in finding unofficial world records of fastest 3x3x3
solves, countries with large proportion of female cubers as well as
acceptable solving times before entering a WCA competition.

<center>
```{r,results='asis',echo=FALSE,fig.cap=""}
cat(paste0("<img src=\"{{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"QUANTILEAVG-1.png\" width=\"550\">\n"))
```
</center>


{% include license.html %}


## Motivation

An item on my bucket list of
irrelevant-things-to-excel-in-during-parental-leave™ was to solve
the Rubik's cube again. In 2013 I learned the [**beginner's method**](http://starlingsroost.asuscomm.com/cubestation/index.php?page=3x3x3/beginner/step1)
for solving the cube by reading
[Dan Harris'](https://www.speedsolving.com/wiki/index.php/Dan_Harris)
*Speedsolving the Cube* book. However, any attempts to improve
times by learning [CFOP](https://en.wikipedia.org/wiki/CFOP_Method)
failed miserably in 2013 due to lack of time. So in this 2019 attempt
I just try to revoke my beginner's method and improve it slightly by
[4LL](https://www.cubeskills.com/tutorials/4-look-last-layer). Nevertheless, my
solves fluctuate at 120-180s. This has made me worry about the
shame-factor of pulling through the next item on the bucket list:
competing in a
[World Cubing Association (WCA)](https://www.worldcubeassociation.org)
3x3x3 competition. The motivating question of this post is
therefore: how large a **skill outlier** would one be at such a competition
with, say, an 180s average?

## Mining the World Cubing Association Database

Luckily, the WCA provides the results of all cubers and competitions
in a
[downloadable SQL database](https://www.worldcubeassociation.org/results/misc/export.html)
(400 MB mySQL/MariaDB SQL text file). The downloadable version of 19 Apr
2019 contains tables with information on all 124,997 players, 6,014
competitions as well as the 2,130,374 results obtained by players in
competitions starting with the [Rubik's Cube World Championship in 1982](https://www.youtube.com/watch?v=m4qoeJMPv0Y). Not **big data**, but big enough and structured in a
way that justifies using a SQL database backend for the analysis
instead of just a flat `data.frame` in R.

With a little bit of effort, one installs a
[MariaDB](https://mariadb.org) together with the
[MariaDB Connector/C](https://downloads.mariadb.org/connector-c/) and
imports the WCA data into the database. Subsequently, the `RMariaDB`
[@rmariadb] and `DBI` [@DBI] R packages allow one to connect directly to the
DB. Furthermore, the `dbplyr` [@dbplyr] package allows one to write
`dplyr` code, which is then translated into and evaluated as SQL
queries behind the scenes.

We connect to the database as follows:
```{r, echo=TRUE}
library(DBI)
library(RMariaDB)
con <- DBI::dbConnect(RMariaDB::MariaDB(), group="my-db")
```
where, as suggested in the [`RMariaDB` package documentation](https://mariadb.com/kb/en/library/configuring-mariadb-with-option-files/), `my-db`
is a custom group defined in the option file `~/my.cnf` containing the
user, password, protocol type and the database to use (`wca`). This is
preferred instead of calling `dbConnect` with options such as `socket`,
`user`, `password` and `dbname`.

To import the data one can either open a `mysql` shell and type SQL
commands or send the SQL query from R to the database:
```{SQL, echo=TRUE, eval=FALSE, results="hide"}
SOURCE WCA_export.sql;
```

```{r, echo=FALSE, results="hide"}
dbExecute(con, "DROP TABLE IF EXISTS Competitions2;")
```
After such an import we can view the contents of the database:
```{r, echo=TRUE}
dbListTables(con)
```
Of particular interest is the `Results` table, where each row contains the
results of a player (`personId`) in a round of a WCA competition of events such as
the 2x2x2, 3x3x3, 4x4x4, etc. For the [3x3x3 event](https://www.worldcubeassociation.org/files/WCA_Competition_Tutorial.pdf) each round consist of five solves
solves (`values1`-`values5`) of which the best is reported in
the column `best` and the trimmed mean (removing the best and worst
solve and computing the mean of the three other solves) is
reported in the column `average`.
```{r}
dbListFields(con, "Results")
```
Furthermore, the table `Persons`contains additional basic information about
each player:
```{r}
dbListFields(con, "Persons")
```
Note: At WCA competitions there is no separate ranking according
to gender.

After getting initially acquainted with the database, we do two small
SQL-side modifications. The first modification is to create an index
for the tables, which will help to substantially reduce the join times
later on. The second is to merge the separate year, month, day columns
of the event day in the `Competitions` table into a single ISO 8601
date. See the [github code](`r paste0("https://raw.githubusercontent.com/hoehleatsu/hoehleatsu.github.io/master/_source/",current_input())`) for details.

```{r, results="hide"}
##Server-side Convert year, month and day column in a proper Date
dbExecute(con, "CREATE TABLE IF NOT EXISTS Competitions2 AS SELECT *, CAST(CONCAT(year, '-', month, '-', day) AS DATE) as date FROM Competitions")

##Create index for faster joining
dbExecute(con, "CREATE INDEX IF NOT EXISTS id_hash ON Persons (id);")
dbExecute(con, "CREATE INDEX IF NOT EXISTS id_hash ON Competitions2 (id);")
dbExecute(con, "CREATE INDEX IF NOT EXISTS id_hash_p ON Results (personId);")
dbExecute(con, "CREATE INDEX IF NOT EXISTS id_hash_c ON Results (competitionId);")
```

As the next step we create tables to use with `dbplyr`. These will then
just dispatch to the database tables. For now, we work only with the
results of the 3x3x3 cubes, but will join the Results, Persons and
Competitions table in order to get relevant information about gender
of the person as well as date of the solve together with the time of
each solve.

```{r, echo=TRUE}
suppressPackageStartupMessages(library(dbplyr))
suppressPackageStartupMessages(library(dplyr))

results <- tbl(con, "Results") %>% filter(eventId == "333")
persons <- tbl(con, "Persons")
competitions <- tbl(con, "Competitions2")

##JOIN the three tables together
allResults <- results %>% inner_join(persons, by=c("personId"="id")) %>%
  inner_join(competitions, by=c("competitionId"="id"))
```

Note that `dbplyr` uses lazy evaluation, i.e. calls are not executed
before needed. However, one can with `show_query` check the SQL call,
which is used in case of evaluation. In this example the SQL query to get
the results of female cubers, i.e.:
```{r, echo=TRUE}
fwr_single <- allResults %>% filter(gender == "f", best>0)
```
in SQL looks like
```{r, echo=TRUE}
fwr_single %>% show_query()
```

## Analysing the data

We are now all ready to perform some descriptive analyses of the
data. The top-5 females according to their time for 3x3x3 single solve
is obtained by:
```{r, echo=TRUE}
fwr_single %>%
  group_by(personId) %>% top_n(-1, best) %>% ungroup %>%
  arrange(best) %>%
  select(competitionId, personName, personCountryId, best, date) %>%
  top_n(-5, best)
```
Note that the solve times are given in centiseconds (i.e. 1/100'th of a second).
The above result can be compare with the corresponding information at
[https://wcadb.net](https://wcadb.net/rankings.php?region=World&events=333&gender=female&show=100&type=single).

As professional data scientist one knows how important it is to
understand the data generating process and to know your data. So this
is how the 5.37s solve by Dana Yi in 2017 looks like:
<p>
<center>
<iframe width="560" height="315"
src="https://www.youtube.com/embed/WMd6JgC4DoQ" frameborder="0"
allow="accelerometer; autoplay; encrypted-media; gyroscope;
picture-in-picture" allowfullscreen></iframe>
</center>
<p>

### The evolution of the 3x3x3 single solve WR

```{r, echo=FALSE, results="hide"}
##Function to compute the indicator function, whether the sequential relative rank
##of a value is equal to one. Note: index of vectors start at 0 in Rcpp.
Rcpp::cppFunction('NumericVector sequential_is_rank1(NumericVector x) {
  int n = x.size();
  NumericVector output(n);
  double best = 1e99;

  for (int i = 0; i < n; ++i) {
     output[i] = x[i] < best;
     if (x[i] < best) { best = x[i]; }
  }
  return output;
}')
```

So at this point we force an execution of the `allResults` query and
cache the result as an object in R.  This feels slightly
disappointing, because the hope was to leave the data in the database
management system (DBMS) as long as possible, but it felt like the
most efficient way to compute sequential ranks -- however, it might
have been possible to perform the sequential rank directly using SQL
statements, although I did not succeed to find the correct approach
within the available time [^1].

```{r}
allResultsTab <- allResults %>% collect()
```

Instead, the results are sorted according to their date in R and
subsequently each result is checked to see if it's sequential rank is
1, i.e. whether the time is lower than all previous results.  For this
purpose a fast Rcpp function `sequential_is_rank1` function is provided,
which computes the sequential rank of a vector of values. Note:
If we had not pulled the data into R at this point, such a computation
within R would not have been possible.

```{r, warning=FALSE}
######################################################################
## Extract all gender specific new WRs in the period
## [from_year-01-01, 2019-04-19]
##
## @param which_gender Gender to consider (either "m" or "f")
## @param from_year Start of the time period to report results for
## @return A data.frame ordered by `date` where each row corresponds
##         to a result which was a new 3x3x3 single WR among `gender`.
######################################################################
wr_evolution <- function(which_gender, from_year=2005) {
  single <- allResultsTab %>%
    filter(gender == which_gender, best>0) %>% arrange(date)

  ##Compute the sequential ranks and select only new WRs in a given time period (>= from_year)
  single_rr <- single %>%
    mutate(rank=sequential_is_rank1(best)) %>%
    filter(rank == 1,  lubridate::year(date) >= from_year)

  ##Add an extra data point at the end allowing geom_step to continue drawing the line
  df <- data.frame(personId="NA", date=as.Date("2019-04-19"), best=min(single_rr$best), rank=1, gender=which_gender)
  res <- full_join(single_rr, df, by=c("personId","date","best","rank", "gender"))
  return(res)
}

##Compute evolution for both genders and combine into one data.frame
wr <- purrr::map_df(c("f","m"), wr_evolution)
```

The evolution of the world record over time is then easily plotted:
```{r WREVO, message=FALSE, }
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(lubridate))
ggplot(wr, aes(x=date, y=best/100, color=gender)) + geom_step() +
  ylab("Single solve 3x3 x3 (s)") +
  xlab("Date of result") +
  scale_color_viridis(discrete=TRUE) +
  scale_x_date(breaks = seq(as.Date("2005-01-01"), as.Date("2019-04-19"), by = "2 year"), labels=year(seq(as.Date("2005-01-01"), as.Date("2019-04-19"), by = "2 year"))) +
  scale_y_continuous(limit=c(0,NA))
```
Useful? Not really IMHO, but I had seen a similar plot at the
[WCA webpage](https://www.worldcubeassociation.org/results/misc/evolution/),
so it's nice to be able to create it in R.

### Countries with highest proportion of female cubers

Since the overall fraction of female cubers is around 10%, we
determine the top-5 countries (with at least 50 cubers), having the
highest proportion of female cubers in the `Persons` database:

```{r, message=FALSE, warning=FALSE}
persons %>% group_by(id) %>% group_by(countryId) %>%
  summarise(n_total=n(), n_male=sum(gender=="m"), n_female=sum(gender=="f")) %>%
  mutate(`frac_female (in %)`= n_female/n_total*100)  %>% ungroup %>%
  collect() %>% ##collect nessary, otherwise it doesn't work?
  filter(n_total >= 50) %>%
  top_n(5, `frac_female (in %)`) %>%
  kable(digits=1, format = "html") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```
<p>
I find this list quite surprising and also encouraging!

### Skill level before entering a WCA competition

Getting back to the motivating question of this post, we derive for
each result how
experienced the cuber was at the time of obtaining the result. We
shall measure experience terms of years since the cuber's first WCA
competition. Special interest will then be in results of cubers in their
very first WCA competition - see the [github code](`r paste0("https://raw.githubusercontent.com/hoehleatsu/hoehleatsu.github.io/master/_source/",current_input())`) for details.

```{r, message=FALSE, warning=FALSE}
##Find first WCA competition of each cuber
first <- allResultsTab %>% group_by(personId) %>% select(date) %>%
  summarise(first_wca_competition = min(date))
##Form an "experience" column containing the number of years since the
##first WCA competition
allResultsTab2 <- inner_join(allResultsTab, first, by="personId") %>%
  mutate(experience=as.numeric((date - first_wca_competition)/365.25),
         resultYear=year(date))
##Look only at results with a valid average (i.e. 5 results,
##where the best and worst result are removed and the remaining 3 averaged)
allResultsTabAvg <- allResultsTab2 %>% filter(average>0)

##We will only analyse results of the last year and we will remove
#the WCA 1982 championship participants
res_lastyear <- allResultsTabAvg %>%
  filter(date >= as.Date("2018-04-19"), year(first_wca_competition)>=2003)
```
```{r}
##Quantiles to consider
q_grid <- c(0.05, 0.1, 0.5, 0.9, 0.95)
```
We then use this information to create a scatter plot with result
average (in seconds)
and corresponding experience of the cuber (in years). For better
visualization we plot the marginal
`r paste0(sprintf("%0.f%%",q_grid*100), collapse=", ")`
quantiles as smooth function of experience - this is done with
with `ggplot2`'s  `geom_quantile` function together with the argument
`method="rqss"`, which then uses the `rqss` function of the `quantreg`
package [@quantreg] to compute smooth quantile curves:

```{r QUANTILEAVG, message=FALSE, warning=FALSE}
pal <- RColorBrewer::brewer.pal(n=3, name="Set2")
##Plot quantiles
ggplot(res_lastyear, aes(x=experience, y=average/100)) + geom_point(color=pal[1], alpha=0.05) +
  geom_quantile(method = "rqss", lambda = 0.1, quantiles=q_grid, color=pal[2]) +
  coord_cartesian(ylim=c(0,60), xlim=c(0,10)) +
  xlab("Experience (years)") + ylab("Average 3x3x3 (s)")
```
We notice that the quantile curves stay more or less parallel, which
is indicative of a stable variance and skewness of the results over the range of
experiences. Focusing only on the round-1 results of those
participating for the first time in the period
from 2018-04-19 to 2019-04-19 we see that the quantiles for the average is
(in seconds) are:
```{r}
res_lastyear_newbies <- res_lastyear %>% filter(experience==0, roundTypeId==1)
```

```{r, eval=FALSE}
ggplot(res_lastyear_newbies, aes(x=average/100, y=..count../sum(..count..))) + geom_histogram(breaks=seq(0,180,by=5), fill=pal[1], alpha=0.8) + xlab("Average 3x3x3 (s)") + ylab("Proportion of results") + scale_y_continuous(labels=scales::percent) + coord_cartesian(xlim=c(0,180)) + scale_color_viridis()
```

```{r}
q <- quantile(res_lastyear_newbies$average/100, probs=c(q_grid, 0.99, 0.999))
print(q, digits=1)
```
This shows that with a 180s average one is located at
the `r sprintf("%.1f%%",100*mean(res_lastyear_newbies$average<=18000))`
quantile of all cubers entering a WCA competition. In other words: if the
comfort zone is defined as **being within the 95%
envelope**, then a ~90s average is needed before entering
a WCA competition.

To further investigate, how cubers of that skill level evolve in time, we study the
solving skills of cubers entering their first WCA competition with a
solve time between 180s and 240s. In order to reduce the
potential effect of secular trends due to, e.g., better cubes, we
consider the skill evolution of the cohort of **first time cubers** from
2015 and onwards.

```{r COHORTEVOLUTION}
##Define cohort of all cubers entering a WCA competition for the first time
cohort <- allResultsTabAvg %>% filter(year(first_wca_competition) >= 2015) %>%
  filter(experience==0)

##Extract only those with an average time in my league
cohort_myleague <- cohort %>% group_by(personId) %>%
  summarise(best_average=min(average)) %>%
    filter(best_average > 3*6000 & best_average < 4*6000)

cohort_evolution <- allResultsTabAvg %>%
  inner_join(cohort_myleague, by="personId")
```

```{r, results="hide"}
ptab <- cohort_evolution %>% group_by(personId) %>%
  distinct(competitionId) %>%
  summarise(n=n()) %$% n %>% table() %>% prop.table()
structure(sprintf("%.1f%%",ptab*100), names=names(ptab))
```

The cohort inclusion criterion provide a total of `r cohort_myleague %>% nrow()` first time
competitors in this skill bracket. Only `r sprintf("%.1f%%",100*(1- sum(ptab[1])))` of these cubers decide
to participate in further WCA competitions! The further development of
the averages of these cubers is best shown in a trajectory
plot. Note that the end of the lines does not necessarily mean that
they stopped cubing, instead it could be due to right truncation,
because only competitions until 2019-04-19 are available.

```{r TRAJPLOT, message=FALSE}
ggplot(cohort_evolution, aes(x=experience, y=average/100, color=as.factor(personId))) + geom_line() + geom_point() +
#  scale_color_viridis(discrete=TRUE, guide=FALSE) +
  scale_color_discrete(guide=FALSE) +
  scale_y_continuous(limits=c(0,NA)) +
  scale_x_sqrt() +
  ylab("Average 3x3x3 (s)") + xlab("Years of experience") +
  geom_hline(yintercept=3*6000/100, lty=2, col="lightgray") +
  geom_hline(yintercept=4*6000/100, lty=2, col="lightgray")
```

Instead of the trajectories we can also overlay an expectation
smoother on top of the data to see how the expected average progresses
with time in the cohort. Note, that this portrays the marginal
expectation and thus is only based on cubers, who are still cubing at
that time. No adjustment for any, potentially informative, drop-out is
made.

```{r TRAJSMOOTHED, warning=FALSE, message=FALSE}
ggplot(cohort_evolution, aes(x=experience, y=average/100)) +
  geom_line(aes(color=as.factor(personId)), alpha=0.2) +
  geom_point(aes(color=as.factor(personId)),alpha=0.2) +
  scale_color_discrete(guide=FALSE) +
  geom_smooth() +
  scale_y_continuous(limits=c(0,NA)) +
  geom_hline(yintercept=3*6000/100, lty=2, col="lightgray") +
  geom_hline(yintercept=4*6000/100, lty=2, col="lightgray") +
  ylab("Average 3x3x3 (s)") + xlab("Years of experience")
```

From the figure we notice a rapid improvement the first 6 months
after entering the first WCA competition. Hereafter results only
improve slowly.

## Discussion

Through analysis of the WCA results database it became clear that
participating in a 3x3x3 event with a 180s average is uncool. The data
also show that cubers entering the world of WCA competitions with such
an average are likely to never participate in another WCA event. In
case they do, their times drop to 90-120s averages within 6 months
after which they are stuck - no one cracks the 20s barrier. To
conclude: It seems wise to practice more, before going to the first
WCA competition. `r emo::ji("smiley")`

From a data science perspective this post provided insights on using
`dbplyr` as a tidyverse frontend for SQL queries. Being an SQL novice,
I learned that generating indexes is a *must*, if you do not want to
wait forever for your *INNER JOIN*s. Furthermore, I ran into some
trouble with mutates and filters with the package, because the
produced SQL code provide empty results. It remains unclear to me if
this was due to differences in SQL DBMSs (e.g. MariaDB being
differnet from PostGres), my lack of SQL knowledge or if it's a shortcoming
of the `dbplyr` package. My conclusion is that performing the data
wrangling within the DBMS was overkill for the medium sized wca data. For larger or more structured datasets such an approach can,
however, be fruitful, because a DBMS's main purpose is to provide
efficient solutions for working with your data. Furthermore, I like
the idea of having a familiar dplyr frontend and just auto-generate an
efficient backend code (here: SQL) to do the actual wrangling. This
also provides an opportunity to get an outside-R-implementation, which
can be an important aspect in
[quality assurance of statistical analyses](http://staff.math.su.se/hoehle/blog/2017/09/02/pairprogramming.html).



## Acknowledgments

The terms of use of the WCA database requests any use of it to be
equipped with the following text:

> This information is based on competition results owned and maintained by the
> World Cube Association, published at https://worldcubeassociation.org/results
> as of April 19, 2019.

Besides this formal note, I thank the WCA Results Team for providing
the WCA data for download in this comprehensive form!

[^1]: If you know how to do this efficiently in SQL, let me know!

## Literature
