---
layout: post
title: "The Mathematics and Statistics of Infectious Disease Outbreaks"
tags: [rstats, COVID-19, SARS-CoV-2, mathematical modelling]
#  bibliography: ~/Literature/Bibtex/jabref.bib
header-includes:
   - \usepackage{bm}
comments: true
editor_options:
  chunk_output_type: console
---

```{r,include=FALSE,echo=TRUE,message=FALSE}
##If default fig.path, then set it.
if (knitr::opts_chunk$get("fig.path") == "figure/") {
  knitr::opts_knit$set( base.dir = '/Users/hoehle/Sandbox/Blog/')
  knitr::opts_chunk$set(fig.path="figure/source/2020-11-20-infepi/")
}
fullFigPath <- paste0(knitr::opts_knit$get("base.dir"),knitr::opts_chunk$get("fig.path"))
filePath <- file.path("","Users","hoehle","Sandbox", "Blog", "figure", "source", "2020-11-20-infepi")

knitr::opts_chunk$set(echo = TRUE,fig.width=8,fig.height=4,fig.cap='',fig.align='center',echo=TRUE,dpi=72*2)#, global.par = TRUE)
options(width=150, scipen=1e3)

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(viridis))

# Non CRAN packages
# devtools::install_github("hadley/emo")

# Fix seed
set.seed(42)

##Configuration
options(knitr.table.format = "html")
theme_set(theme_minimal())
#if there are more than n rows in the tibble, print only the first m rows.
options(tibble.print_max = 10, tibble.print_min = 5)
```

## Abstract:

Slides, R code and video lectures of our 2020 *The Mathematics and Statistics of Infectious Disease Outbreaks* summer course at Stockholm University are made available to a wider audience.

<p>
<center>
```{r,results='asis',echo=FALSE,fig.cap=""}
cat(paste0("<img src=\"{{ site.baseurl }}/",knitr::opts_chunk$get("fig.path"),"youtube.png\" width=\"550\">\n"))
```
</center>

{% include license.html %}

## Introduction

During the 2020 summer  [Tom Britton](https://staff.math.su.se/tom.britton/) and I gave a course on *The Mathematics and Statistics of Infectious Disease Outbreaks* at the [Department of Mathematics, Stockholm University, Sweden](https://www.math.su.se). Pre-requisites for the course were undergraduate knowledge of mathematics (e.g. differential equations, optimization) and statistics (e.g. random variables, distributions, maximum likelihood inference) as well as some programming skills in a language with a data science component (python, R, Julia, matlab, ...).

Now the course is done, we have decided to share all our course material, consisting of slides, R code and video lectures. The main page for navigating the material is on GitHub:
<p>
<center>
[https://github.com/hoehleatsu/mt3002-summer2020](https://github.com/hoehleatsu/mt3002-summer2020)
</center>
<p>

which, e.g., links to the [Youtube playlist](https://www.youtube.com/playlist?list=PLl_ncesshp_C-URStf-LwhYCbFGE70_jq) containing the videos. 

## Course content

* Introduction to the Course [[Tom Britton](https://video.su.se/media/Introduction+to+the+Course+%28Tom+Britton%29/0_scrd1q6o/356471) | [Michael Höhle](https://video.su.se/media/t/0_rqwxd9zg)]
* L01: Mathematical Modelling [[Part 1](https://video.su.se/media/L01+-+Mathematical+Modelling+%281+2%29/0_g9pf4rnh/356471) | [Part 2](https://video.su.se/media/L01+-+Mathematical+Modelling+%282+2%29/0_qdurs7hj)]
* L02: Simulation and Fitting of Epidemic Models [[Part 1](https://video.su.se/media/L02+-+Simulation+and+Fitting+of+Epidemic+Models+%281+3%29/0_j0v2lslj) | [Part 2](https://video.su.se/media/L02+-+Simulation+and+Fitting+of+Epidemic+Models+%282+3%29/0_p4629rt7) | [Part 3](https://video.su.se/media/L02+-+Simulation+and+Fitting+of+Epidemic+Models+%283+3%29/0_3yjyzq0q)]
* L03: Timing and observations + Endemic models [[Part 1](https://video.su.se/media/L03+-+Timing+and+observations+%2B+Endemic+models+%281+2%29/0_10homfzz) | [Part 2](https://video.su.se/media/L03+-+Timing+and+observations+%2B+Endemic+models+%282+2%29/0_1p5i3vsj/356471)]
* L04: Estimating Reproduction Numbers [[Part 1](https://video.su.se/media/L04+-+Estimation+reproduction+numbers+%281+2%29/0_w12jeszk) | [Part 2](https://video.su.se/media/L04+-+Estimating+Reproduction+Numbers+%282+2%29/0_bopogny5)]
* L05: Effective Reproduction Number [[Part 1](https://video.su.se/media/L05+-+Effective+Reproduction+Number+%281+2%29/0_u3b1j7rk) | [Part 2](https://video.su.se/media/L05+-+Effective+Reproduction+Number+%282+2%29/0_38cwcho8)]
* L06: Latencies and Delays [[Part 1](https://video.su.se/media/+L06+-+Latencies+and+Delays+%281+2%29/0_vrzkcn4s) | [Part 2](https://video.su.se/media/L06+-+Latencies+and+Delays+%282+2%29/0_0xu0jgzy)]
* L07: Vaccination, other Preventive Measures and Uncertainties [[Part 1](https://video.su.se/media/L07+-+Vaccination%2C+other+preventive+measures+and+uncertainties+%281+2%29/0_9oson513/356471) | [Part 2](https://video.su.se/media/L07+-+Vaccination%2C+other+preventive+measures+and+uncertainties+%282+2%29/0_7rrq3ug7/356471)]
* L08: Modeling using Networks and other Heterogeneities [[Part 1](https://video.su.se/media/L08+-+Modeling+using+networks+and+other+heterogeneities+%281+2%29/0_y4jqv270/356471) | [Part 2](https://video.su.se/media/L08+-+Modeling+using+networks+and+other+heterogeneities+%282+2%29/0_0mweom9s/356471)]
* L09: Univariate Outbreak Detection [[Part 1](https://video.su.se/media/L09+-+Univariate+outbreak+detection+%281+2%29/0_imgomlwy/356471) | [Part 2](https://video.su.se/media/L09+-+Univariate+outbreak+detection+%282+2%29/0_ifu44eas/356471)]
* L10: Multivariate Outbreak Detection [[Part 1](https://video.su.se/media/L10+-+Multivariate+outbreak+detection+%281+2%29/0_uithqyum) | [Part 2](https://video.su.se/media/L10+-+Multivariate+outbreak+detection+%282+2%29/0_mmol88iy)]
* L11: COVID-19 (I): Reproduction Number and Herd Immunity [[Part 1](https://video.su.se/media/L11+-+COVID-19+%281+2%29/0_dk0ztmz6/356471) | [Part 2](https://video.su.se/media/L11+-+COVID-19+%282+2%29/0_j1xks7xm)]
* L12: COVID-19 (II): Digital Contact Tracing [[Part 1](https://video.su.se/media/L12+-+Digital+Contact+Tracing+%281+2%29/0_9chh75x5) | [Part 2](https://video.su.se/media/L12+-+Digital+Contact+Tracing+%282+2%29/0_pbdfeaov)]

## Discussion

We hope the material can be of value for those interested in the field, e.g., new Ph.D. students in epidemic modelling, infectious disease epidemiologists with a like for the quantitative side of matters, and for those who just want to improve their armchair epidemiology skills.
